---
title: "R in Action: Code, Analyze & Contribute"
subtitle: "Day 4 — RNA-seq Case Study | Participant Workbook"
author: "R User Group Nepal × KU Biotechnology Creatives"
date: "April 12, 2026"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

------------------------------------------------------------------------

## 🧬 RNA-seq Case Study

------------------------------------------------------------------------

> *"Everything you built today with health data — layers, facets, line plots, patchwork — is exactly the toolkit bioinformaticians use with genomics data."*

This case study uses RNA-seq data from **Fu et al. 2015** (GEO: GSE60450). The study examined gene expression in **basal** and **luminal** mammary cells from mice at three stages: virgin, pregnant, and lactating. There are **2 samples per group × 6 groups = 12 samples total**.

### Key RNA-seq Terminology

| Term | Meaning | Health Analogy |
|----|----|----|
| Gene | Unit of DNA coding for a protein | A specific blood test |
| Expression | How actively a gene is used | The test result value |
| Count / CPM | Normalized RNA fragments detected | The measurement reading |
| Sample | One biological specimen | One patient visit |
| Condition | Experimental group | Control vs Treatment |
| Replicate | Repeated measurement | Duplicate readings |
| Differential expression | Gene significantly up/down | Abnormal lab value |
| log2 fold change | Magnitude of change | 2× higher / lower |
| p-value / FDR | Statistical confidence | Same as in any test |

| Group Code | Cell Type | Stage     |
|------------|-----------|-----------|
| `bvirg`    | Basal     | Virgin    |
| `bpreg`    | Basal     | Pregnant  |
| `blact`    | Basal     | Lactating |
| `lvirg`    | Luminal   | Virgin    |
| `lpreg`    | Luminal   | Pregnant  |
| `llact`    | Luminal   | Lactating |

------------------------------------------------------------------------

```{r}
library(tidyverse)
```

------------------------------------------------------------------------

## Steps 1–3: Load, Explore, Reshape

### 🔄 Your Turn CS-01 — Load & Explore *(5 minutes)*

> **Scenario:** Two CSV files are waiting in the `data/` folder. Load them, rename the first column of each, and do a quick exploration.

```{r}
# Load the counts file (CPM and TMM normalized data)
# Filename: GSE60450_GeneLevel_Normalized(CPM.and.TMM)_data.csv
counts <- read_csv("data/___", show_col_types = FALSE)

# Load the sample metadata file
# Filename: GSE60450_filtered_metadata.csv
sampleinfo <- read_csv("data/___", show_col_types = FALSE)
```

```{r}
# Explore sampleinfo — try at least 3 of: head(), dim(), tail(), names(), str()
___(sampleinfo)
___(sampleinfo)
___(sampleinfo)

# Look at the first 3 values of the characteristics column
sampleinfo$___[1:3]
```

```{r}
# Rename the first column of each data frame to a cleaner name
# sampleinfo: first column → "sample_id"
# counts:     first column → "gene_id"
colnames(___)[1] <- "sample_id"
colnames(___)[1] <- "gene_id"
```

```{r}
# Quick summary of counts — what range of values do you see?
summary(___)
```

------------------------------------------------------------------------

### Reshape: wide → long

The counts file has one column per sample. We need to reshape it so every row is one gene × one sample (long format) — the same idea as when you used `pivot_longer()` in the tidyr session.

```{r}
# pivot_longer: collect all sample columns (everything except gene_id and gene_symbol)
# into two new columns: Sample (the old column name) and Count (the value)
seqdata <- pivot_longer(
  counts,
  cols      = -c("___", "___"),   # exclude these two identifier columns
  names_to  = "___",              # new column for old column names
  values_to = "___"               # new column for the expression values
)

# How many rows does seqdata have now?  (expect: n_genes × 12)
nrow(seqdata)
```

------------------------------------------------------------------------

## Steps 4–7: Join, QC, Label Groups

### 🔄 Your Turn CS-02 — Join & QC Plot *(5 minutes)*

> **Scenario:** Combine the expression data with sample metadata, then make a quality-check boxplot to confirm all 12 samples are comparable.

```{r}
# Join seqdata (Sample column) with sampleinfo (sample_id column)
# Use full_join() — same pattern as joining two data frames in dplyr
allinfo <- full_join(
  seqdata,
  sampleinfo,
  by = join_by(___ == ___)   # seqdata column == sampleinfo column
)

allinfo
```

```{r}
# QC boxplot: one box per sample, y-axis = log2(Count + 1)
# We add + 1 before log2 to handle zero counts (log2(0) is undefined)
# Rotate x-axis labels 90° so sample names don't overlap
ggplot(data = allinfo,
       mapping = aes(x = ___, y = log2(___ + 1))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = ___)) +
  labs(title = "Quality check: all samples comparable?",
       y     = "log2(CPM + 1)")
```

```{r}
# Colour the boxes — add fill = Sample inside aes()
# Does this make it easier to see which samples stand out?
ggplot(data = allinfo,
       mapping = aes(x = Sample, y = log2(Count + 1), fill = ___)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-03 — Label the Groups *(5 minutes)*

> **Scenario:** The `characteristics` column contains long strings like `"cell type: basal; developmental stage: virgin"`. Use `case_when()` and `str_detect()` to create short, clean group labels.

```{r}
# Create a new column "Group" using mutate() + case_when()
# str_detect() checks if the characteristics string contains a pattern
# Pattern "basal.*virgin" means: "basal" followed anywhere by "virgin"
allinfo <- mutate(allinfo, Group = case_when(
  str_detect(characteristics, "basal.*virgin")   ~ "___",   # → "bvirg"
  str_detect(characteristics, "basal.*preg")     ~ "___",   # → "bpreg"
  str_detect(characteristics, "basal.*lact")     ~ "___",   # → "blact"
  str_detect(characteristics, "luminal.*virgin") ~ "___",   # → "lvirg"
  str_detect(characteristics, "luminal.*preg")   ~ "___",   # → "lpreg"
  str_detect(characteristics, "luminal.*lact")   ~ "___"    # → "llact"
))

# Check: should be 2 samples per group (2 replicates × n_genes rows each)
table(allinfo$___)
```

------------------------------------------------------------------------

## Steps 8–12: Select Top Genes and Plot

### 🔄 Your Turn CS-04 — Expression Plot *(8 minutes)*

> **Scenario:** Find the 8 most highly expressed genes across all samples, then visualise their expression by group using jitter + facets.

```{r}
# Use a dplyr pipe to:
#   1. group_by gene_symbol
#   2. summarise: total_count = sum(Count)
#   3. arrange by total_count descending
#   4. take the top 8 with head(n = 8)
#   5. pull just the gene_symbol column as a vector
top_genes <- allinfo %>%
  group_by(___) %>%
  summarise(total_count = ___(Count)) %>%
  arrange(___(total_count)) %>%
  head(n = ___) %>%
  pull(___)

top_genes

# Filter allinfo to only the top genes
mygenes_counts <- filter(allinfo, gene_symbol %in% ___)
mygenes_counts
```

```{r}
# Basic jitter plot — one panel per gene (facet_wrap)
# Map colour to Group
ggplot(data = mygenes_counts,
       mapping = aes(x = ___, y = log2(Count + 1), colour = ___)) +
  geom_jitter(size = 3, width = ___) +
  facet_wrap(~ ___) +
  theme_classic()
```

```{r}
# Customise: apply manual colours and proper axis labels
# 6 colours for 6 groups — pick your own if you like!
mycolours <- c("turquoise", "plum", "tomato", "violet", "steelblue", "chocolate")

ggplot(data = mygenes_counts,
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter(size = 3, width = 0.15) +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = ___) +
  labs(x        = "___",
       y        = "Expression (log2 CPM + 1)",
       title    = "Mammary Gland Gene Expression",
       subtitle = "Fu et al. 2015, GSE60450") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text  = element_text(face = "bold"))
```

```{r}
# Fix the x-axis ORDER using factor levels — without this, R sorts alphabetically
# bvirg → bpreg → blact → lvirg → lpreg → llact (developmental order)
group_order <- c("bvirg", "bpreg", "blact", "lvirg", "lpreg", "llact")

mygenes_counts <- mutate(mygenes_counts,
  Group_f = factor(___, levels = ___))

# Check the levels are set correctly
levels(mygenes_counts$___)
```

```{r}
# Final plot using Group_f (ordered factor) on x-axis
ggplot(data = mygenes_counts,
       mapping = aes(x = ___, y = log2(Count + 1), colour = ___)) +
  geom_jitter(size = 3, width = 0.15) +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = mycolours) +
  labs(x        = "Cell Type and Stage",
       y        = "Expression (log2 CPM + 1)",
       title    = "Mammary Gland Gene Expression",
       subtitle = "Fu et al. 2015, GSE60450") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text  = element_text(face = "bold"))
```

```{r}
#| eval: false
# Save the final plot as PNG
ggsave("plots/mammary_expression.png", width = ___, height = ___, dpi = 300)
```

------------------------------------------------------------------------

### ✅ Exercises

**Exercise 1:** Replace `mycolours` with 6 colors of your own choice. Try `scale_fill_brewer(palette = "Set2")` as an alternative.

**Exercise 2:** Change `facet_wrap(~ gene_symbol)` to `facet_wrap(~ gene_symbol, ncol = 2)`. What changes in the layout?

**Exercise 3:** Add `geom_boxplot(alpha = 0.2)` as a layer **underneath** the jitter. Which line do you write first?

**Exercise 4:** Filter `mygenes_counts` to only luminal groups (`lvirg`, `lpreg`, `llact`) and remake the plot.

**Exercise 5:** Add `subtitle = "Data: Fu et al. 2015, GEO: GSE60450"` to your labs and save both as PNG and PDF. (Hint: `pdf("myplot.pdf")` ... `dev.off()`)

**Exercise 6 (★):** Add `geom_hline(yintercept = 8, linetype = "dashed", colour = "gray40")` to the plot. What does this horizontal line represent biologically? *(Hint: log2(CPM + 1) = 8 corresponds to \~255 CPM — what does a high CPM mean?)*

```{r}
# Space for exercises




```

------------------------------------------------------------------------

## 🌋 Volcano Plot — Differential Expression

------------------------------------------------------------------------

### What Is a Volcano Plot?

A volcano plot shows **two statistics simultaneously** for every gene in a differential expression experiment:

-   **X-axis:** `log2 fold change` — direction and size of change (Basal vs Luminal)
-   **Y-axis:** `-log10(adjusted p-value)` — statistical confidence

Genes that are both **large change** (far left or right) AND **statistically significant** (high up) appear in the "eruption" — these are the biologically interesting genes.

| Region | Meaning |
|----|----|
| Top-right | Significantly **up-regulated** in Basal |
| Top-left | Significantly **down-regulated** in Basal (= up in Luminal) |
| Bottom-middle | Changed but not significant — likely noise |
| Far left/right, low y | Large change but not reliable |

### Why `-log10(p-value)`?

```{r}
# p-values become very small — hard to read on a linear scale
# -log10 flips them: larger number = more significant
-log10(c(0.05, 0.01, 0.001, 0.0001))
# →     1.3    2.0   3.0    4.0
# Significance threshold: -log10(0.05) = 1.3
```

### Typical Cutoffs

| Cutoff        | Value                        | Meaning                 |
|---------------|------------------------------|-------------------------|
| Significance  | `p_adjusted < 0.05`          | FDR \< 5%               |
| Fold change   | `abs(log2_fc) > 1`           | At least 2× change      |
| Both combined | Top-right / top-left corners | The "interesting" genes |

------------------------------------------------------------------------

### 🔄 Your Turn CS-05 — Volcano Plot *(8 minutes)*

> **Scenario:** Build the signature figure of differential expression analysis step by step. The `de_results` data has columns `gene_name`, `log2_fc`, `p_value`, and `p_adjusted`.

#### Step 1 — Load DE Results

```{r}
de_results <- read_csv("data/de_results.csv", show_col_types = FALSE)
head(de_results)
glimpse(de_results)
```

#### Step 2 — Add Significance Category

```{r}
# Add two new columns using mutate():
#   neg_log10_p  = -log10(p_adjusted)
#   significance = a category label from case_when()
#
# Cutoffs:  p_adjusted < 0.05  AND  |log2_fc| > 1
de_results <- de_results %>%
  mutate(
    neg_log10_p  = -log10(___),
    significance = case_when(
      p_adjusted < ___ & log2_fc >  ___ ~ "Up in Basal",
      p_adjusted < ___ & log2_fc < -___ ~ "Up in Luminal",
      TRUE                              ~ "Not significant"
    )
  )

# How many genes fall into each category?
de_results %>% count(___)
```

#### Step 3 — Basic Volcano

```{r}
# Scatter plot: x = log2_fc, y = neg_log10_p, colour = significance
ggplot(de_results,
       aes(x = ___, y = ___, colour = ___)) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "___",
               "Not significant" = "gray70")
  ) +
  theme_classic()
```

#### Step 4 — Add Threshold Lines

```{r}
# Add dashed reference lines:
#   horizontal: at -log10(0.05) — significance threshold
#   vertical:   at -1 and +1    — fold change thresholds (two lines at once)
ggplot(de_results,
       aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_hline(yintercept = -log10(___), linetype = "dashed", colour = "gray40") +
  geom_vline(xintercept = c(___, ___),  linetype = "dashed", colour = "gray40") +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "steelblue",
               "Not significant" = "gray70")
  ) +
  theme_classic()
```

#### Step 5 ★ — Label Top Genes & Save

```{r}
# Select the 10 most significant genes (highest neg_log10_p) among significant ones
# Hint: filter() then slice_max(neg_log10_p, n = 10)
top_sig <- de_results %>%
  filter(significance != "___") %>%
  slice_max(___, n = ___)

# Add geom_text() using top_sig as its own data source
# nudge_y moves labels slightly above the points
ggplot(de_results,
       aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "gray40") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed", colour = "gray40") +
  geom_text(data = top_sig,
            aes(label = ___),
            size = 2.8, nudge_y = 0.25, colour = "black") +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "steelblue",
               "Not significant" = "gray70")
  ) +
  labs(x        = "log2 Fold Change (Basal vs Luminal)",
       y        = "-log10(adjusted p-value)",
       title    = "Differential Expression: Basal vs Luminal Cells",
       subtitle = "Fu et al. 2015 · |log2FC| > 1, FDR < 0.05",
       colour   = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

```{r}
#| eval: false
ggsave("plots/volcano_basal_vs_luminal.png", width = 8, height = 6, dpi = 300)
```

------------------------------------------------------------------------

## 📎 Full ggplot2 Quick Reference

``` r
#| eval: false

# ── TEMPLATE ────────────────────────────────────────────────────────────────
ggplot(data, aes(x, y, colour, fill, shape, size, alpha)) +
  geom_*() +
  scale_*() +
  theme_classic() +
  labs(title, subtitle, x, y, colour, caption) +
  ggsave("plots/file.png", width = 8, height = 5, dpi = 300)

# ── GEOMS ───────────────────────────────────────────────────────────────────
geom_point(size = 3, alpha = 0.8)
geom_jitter(width = 0.15, size = 2.5)
geom_line(linewidth = 1)
geom_boxplot(fill = "steelblue", alpha = 0.4)
geom_histogram(bins = 8, fill = "steelblue")
geom_hline(yintercept = 8, linetype = "dashed", colour = "gray40")
geom_vline(xintercept = c(-1, 1), linetype = "dashed")
geom_text(aes(label = gene_name), nudge_y = 0.25, size = 2.8)
annotate("text", x = 130, y = 2.5, label = "Threshold", size = 3)

# ── SCALES ──────────────────────────────────────────────────────────────────
scale_colour_manual(values = c("A" = "tomato", "B" = "steelblue"))
scale_fill_brewer(palette = "Set2")
scale_colour_viridis_c()

# ── FACETS ──────────────────────────────────────────────────────────────────
facet_wrap(~ gene_symbol)
facet_wrap(~ gene_symbol, ncol = 2, scales = "free_y")
facet_grid(gender ~ district)

# ── FACTOR ORDERING ─────────────────────────────────────────────────────────
group_order <- c("bvirg", "bpreg", "blact", "lvirg", "lpreg", "llact")
mutate(data, Group_f = factor(Group, levels = group_order))

# ── VOLCANO ESSENTIALS ───────────────────────────────────────────────────────
-log10(p_adjusted)               # y-axis transform
case_when(...)                   # significance labelling
slice_max(neg_log10_p, n = 10)   # top genes

# ── SAVE ────────────────────────────────────────────────────────────────────
ggsave("plots/file.png", width = 10, height = 8, dpi = 300)
pdf("myplot.pdf"); print(p); dev.off()
```
