---
title: "R in Action: Code, Analyze & Contribute"
subtitle: "Day 1 — Getting Started with R | Participant Workbook"
author: "R User Group Nepal"
date: "April 9, 2026"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

# DAY 1 — Thursday, April 9

**R User Group Nepal × Kathmandu University Biotechnology Creatives** *Online · 8:00 PM – 10:00 PM (Nepal Standard Time)*

------------------------------------------------------------------------

## 👋 Welcome & About R User Group Nepal

**R User Group Nepal (RUG Nepal)** is a community of students, researchers, and professionals who learn and use R together across Nepal.

-   Makes data analysis accessible to Nepali students and researchers
-   Organizes workshops, seminars, and open-source events like this one
-   Connected to the global R community through the R Consortium
-   **You are now part of this community** — welcome!

> *"You don't need to be an expert to join. You just need to be curious."*

------------------------------------------------------------------------

## 🧑‍🏫 Know Your Instructor

You can connect with him over: <https://linkedin.com/in/bjungbogati>

------------------------------------------------------------------------

## 🌍 Why R? Where Is It Actually Used?

R is used every day in real science and research:

| Field                 | What R Does There                                |
|-----------------------|--------------------------------------------------|
| **Biotechnology**     | Analyze lab measurements, patient test results   |
| **Bioinformatics**    | Process genomic and health data                  |
| **Public Health**     | Track disease outbreaks, analyze patient records |
| **Pharmaceuticals**   | Clinical trial statistics, drug analysis         |
| **Agriculture**       | Crop and environmental data analysis             |
| **Academic Research** | Write fully reproducible papers and reports      |

Organizations actively using R:

-   **Pfizer, Novartis, Roche** — clinical trial analysis
-   **WHO, CDC** — disease surveillance and epidemiology
-   **Universities worldwide** — research and publishing

> 💡 R is not just a student tool. It runs in real labs, hospitals, and companies — including research affecting health outcomes in Nepal.

------------------------------------------------------------------------

## 📌 Why R for YOU?

You may already use Excel in your courses. That is fine for small tasks. But consider this:

**Situation:** You collect **vital signs** (blood pressure, heart rate, glucose) for 50 patients. Your supervisor asks you to:

1.  Calculate average blood pressure per age group
2.  Flag patients with high risk (BP \> 140 mmHg)
3.  Make a comparison chart for the report
4.  Redo everything when 5 records are removed

|                          | Excel        | R                      |
|--------------------------|--------------|------------------------|
| Time to complete         | \~45 minutes | \~2 minutes            |
| Risk of click errors     | High         | Very low               |
| Redo after change        | Start over   | Run the script again   |
| Shareable & reproducible | Difficult    | Yes — share the script |

------------------------------------------------------------------------

## ⚙️ Setup: Install R and RStudio

> **If you already installed both, skip ahead to "Let's Look at RStudio."**

**Step 1 — Install R:** <https://cran.r-project.org>

**Step 2 — Install RStudio:** <https://posit.co/download/rstudio-desktop>

> ✅ Open RStudio. If you see 3 or 4 panels — you are ready.

------------------------------------------------------------------------

## 🖥️ Let's Look at RStudio

RStudio has **4 panes**:

```         
┌───────────────────────┬──────────────────────┐
│   Source / Script     │    Environment       │
│   (write code here)   │  (your variables)    │
├───────────────────────┼──────────────────────┤
│   Console             │   Files / Plots /    │
│   (output appears)    │   Packages / Help    │
└───────────────────────┴──────────────────────┘
```

| Pane | What You Use It For |
|----|----|
| **Source / Script** (top-left) | Write and save your code |
| **Console** (bottom-left) | See output after running code |
| **Environment** (top-right) | All variables and datasets you created |
| **Files / Plots / Help** (bottom-right) | Browse files, view graphs, read help |

------------------------------------------------------------------------

## ✍️ Writing Your First Code

Open a Quarto document: `File → New File → Quarto Document` → save as `first_code.qmd`

Press **Ctrl + Enter** (Windows) or **Cmd + Enter** (Mac) to run each line:

```{r}
# The # symbol starts a comment — R ignores it.
# Always comment your code!

# R as a calculator
2 + 3
10 / 2
3 ^ 2

# A health example:
# A nurse recorded systolic blood pressure for 3 patients.
# Systolic BP is the top number in a blood pressure reading (e.g., 118 in 118/78).
# Let's calculate the average:


(118 + 125 + 122) / 3
```

------------------------------------------------------------------------

## 📦 Install and Load Packages

R comes with basic built-in functions. **Packages** add extra tools — like installing apps on a phone.

```{r}
# Install a package — do this ONCE (like downloading an app)
# install.packages("tidyverse")

# Load a package — do this EVERY session (like opening the app)
library(tidyverse)
```

**What is `tidyverse`?** A collection of packages for loading, cleaning, analyzing, and visualizing data:

| Package   | What It Does                     |
|-----------|----------------------------------|
| `readr`   | Read CSV and data files          |
| `dplyr`   | Filter, sort, and summarize data |
| `tidyr`   | Reshape data                     |
| `ggplot2` | Create graphs and plots          |
| `stringr` | Work with text                   |

------------------------------------------------------------------------

## 📦 Create a Project in RStudio

1.  Click `File` → `New Project`
2.  Select `New Directory` → `New Project`
3.  Name your project: **`r_training`**
4.  Click `Create Project`

------------------------------------------------------------------------

## 📊 Your First Plot

```{r}
# Load a built-in dataset about iris flowers
data(iris)

# What does it look like?
head(iris)

# Make a scatter plot
plot(iris$Sepal.Length, iris$Petal.Length,
     main = "Sepal Length vs Petal Length",
     xlab = "Sepal Length (cm)",
     ylab = "Petal Length (cm)",
     col  = "steelblue",
     pch  = 16)
```

> You just created a scientific scatter plot with 4 lines of code!

------------------------------------------------------------------------

# Intro to R Programming

Now let us learn how R actually thinks, so you can write code with confidence.

------------------------------------------------------------------------

## 📋 Things to Remember

-   R is **case-sensitive**: `Patient` and `patient` are different
-   Code runs **line by line** from top to bottom
-   **Errors are normal** — they are messages, not failures
-   The `#` starts a comment — write comments generously

------------------------------------------------------------------------

## 🚫 Reserved Words

Some words belong to R — you cannot use them as variable names:

```         
TRUE    FALSE    NULL    NA    Inf    NaN
if      else     for     while  function  return
```

```{r}
#| error: true
# ❌ This will cause an error — TRUE is reserved
TRUE <- 5
```

```{r}
# ✅ This is fine
is_hypertensive <- TRUE
is_hypertensive
```

------------------------------------------------------------------------

## 📦 Variables

A **variable** is a named container that holds a value. You create one using `<-`:

```{r}
# Storing patient vital signs
patient_name   <- "Sumana"    # text
systolic_bp    <- 118               # number (mmHg)
is_diabetic    <- FALSE             # yes/no
num_patients   <- 50L               # whole number (integer)

# Print a variable — just type its name
patient_name
systolic_bp
```

You can update a variable by assigning a new value:

```{r}
systolic_bp <- 118    # first measurement
systolic_bp

systolic_bp <- 122    # re-checked
systolic_bp
```

R also has built-in constants:

```{r}
pi      # 3.14159...
NA      # missing value
Inf     # infinity
```

------------------------------------------------------------------------

## 🔄 Your Turn 01 — Variables

*(2 minutes)*

```{r}
# 1. Store your own name
my_name <- ""
my_name

# 2. Store a systolic blood pressure reading (in mmHg) — try 118
systolic_bp <- 
systolic_bp

# 3. Store whether a patient has diabetes (TRUE or FALSE)
is_diabetic <- 
is_diabetic

# 4. What happens when you type Systolic_bp instead of systolic_bp?
#    Try it and read the error message carefully.
Systolic_bp
```

> 🔍 **Discussion:** What did the error say? Can you explain what went wrong?

------------------------------------------------------------------------

## 🏷️ Naming Convention

In R, the recommended style is **snake_case** — all lowercase with underscores:

| ✅ Good name    | ❌ Avoid       | Why to avoid               |
|-----------------|----------------|----------------------------|
| `systolic_bp`   | `SystolicBP`   | Hard to read               |
| `patient_count` | `patientcount` | No separation              |
| `is_diabetic`   | `is diabetic`  | Spaces cause errors        |
| `heart_rate`    | `1heart_rate`  | Cannot start with a number |
| `blood_glucose` | `bg`           | Too short — unclear        |

> **Rule of thumb:** Would a classmate understand your variable name without explanation?

------------------------------------------------------------------------

## 🔢 Data Types

R stores different kinds of values differently:

| Type        | What It Stores   | Health Example                                |
|-------------|------------------|-----------------------------------------------|
| `numeric`   | Decimal numbers  | Blood pressure: `118.5`, glucose: `98.0`      |
| `integer`   | Whole numbers    | 50 patients: `50L`, 12 visits: `12L`          |
| `character` | Text (in quotes) | Patient name: `"Avishek"`, status: `"Normal"` |
| `logical`   | TRUE or FALSE    | Is diabetic: `TRUE`, Is smoker: `FALSE`       |
| `factor`    | Categories       | Risk: `"Low"`, `"Medium"`, `"High"`           |

```{r}
# Check what type a value is
class(118.5)       # "numeric"
class("Avishek")     # "character"
class(TRUE)        # "logical"
class(50L)         # "integer"

systolic_bp <- 118
class(systolic_bp) # "numeric"
```

------------------------------------------------------------------------

## 🔄 Your Turn 02 — Data Types

*(2 minutes)*

```{r}
# Guess the type first, then check with class()
class(98.6)
class("Normal")
class(FALSE)
class(10L)
class("10")       # Is this different from 10? Why?
```

```{r}
# Create these variables and check their types
patient_name   <- #"Asia"
num_visits     <- #5L
is_healthy      <- #TRUE
bmi_value      <- #20.5

class(patient_name)
class(num_visits)
class(is_healthy)
```

> 🔍 **Key question:** `class("10")` vs `class(10)` — what is different? Why might it matter in data analysis?

------------------------------------------------------------------------

## 🗂️ Data Structures

How R organizes multiple values:

| Structure | What It Is | When You Use It |
|----|----|----|
| **Vector** | A sequence of values, same type | 6 BP readings from an experiment |
| **Data Frame** | A table — rows and columns | Your full patient dataset |
| **List** | Can hold mixed types | Complex analysis results |

> For Days 1–2, we focus on **vectors** and **data frames**.

------------------------------------------------------------------------

## 📐 Vectors

A **vector** is a sequence of values, all the same type. Created with `c()` (combine):

```{r}
# Systolic blood pressure readings for 6 patients (mmHg)
systolic_bp <- c(118, 125, 132, 145, 110, 138)

# Patient names
patient_names <- c("Avishek", "Bikash", "Chitra",
                   "Dev", "Elina", "Farhan")

# Patient group
groups <- c("Control", "Control", "Control",
            "High Risk", "High Risk", "High Risk")
```

**Working with vectors:**

```{r}
length(systolic_bp)  # how many patients?
mean(systolic_bp)    # average BP
max(systolic_bp)     # highest reading
min(systolic_bp)     # lowest reading
```

**Accessing specific values** — use square brackets `[]`:

```{r}
systolic_bp[1]         # first patient
systolic_bp[3]         # third patient
systolic_bp[1:3]       # first three
systolic_bp[c(1, 4)]   # first and fourth
```

> 💡 Patients 4–6 are in the High Risk group. Patient 4 (145 mmHg) already stands out as having elevated blood pressure — you can see this just by looking at the vector.

------------------------------------------------------------------------

## 🔄 Your Turn 03 — Vectors

*(3 minutes)*

Create a vector called `heart_rate` with the following values:

``` r
72, 85, 68, 91, 78, 88, 74, 95
```

```{r}
# A nurse recorded resting heart rate (bpm) for 8 patients.
# Patients 1–4: standard care group
# Patients 5–8: intensive monitoring group

# values of hear_rate = 72, 85, 68, 91, 78, 88, 74, 95

heart_rate <- 

# 1. How many patients?
length()

# 2. Average heart rate across all patients?
mean()

# 3. Highest and lowest reading?
max()
min()

# 4. Standard care group (first 4 patients)
heart_rate[1:4]

# 5. Intensive monitoring group (last 4 patients)
heart_rate[5:8]

# 6. Average for each group separately
mean(heart_rate[1:])
mean(heart_rate[:8])

# 7. Is the intensive monitoring group average higher?
mean(heart_rate[5:8]) > mean(heart_rate[1:4])
```

> 🔍 **What do you observe?** Which group has a higher average heart rate? What might that mean?

------------------------------------------------------------------------

## ➕ Operators

**Arithmetic Operators:**

```{r}
10 + 3     # addition
10 - 3     # subtraction
10 * 3     # multiplication
10 / 3     # division
10 ^ 2     # power
10 %% 3    # remainder
```

**Relational Operators** — always return TRUE or FALSE:

```{r}
118 < 140     # Is this BP below high threshold?
155 > 140     # Is this BP above high threshold?
118 == 118    # Are these equal? (double ==)
118 != 140    # Are these NOT equal?
118 >= 110    # Is 118 >= 110?
```

> ⚠️ `=` assigns values. `==` compares values. These are different!

**Logical Operators:**

```{r}
# AND — both conditions must be true
118 > 90 & 118 < 140    # Is BP in the normal range?

# OR — at least one must be true
165 < 70 | 165 > 160    # Is BP dangerously out of range?

# NOT — reverses the result
!TRUE
!FALSE
```

**Practical example — flag a patient:**

```{r}
bp <- 165

# Flag if BP is dangerously low OR dangerously high
bp < 70 | bp > 160    # TRUE — this patient needs urgent attention
```

------------------------------------------------------------------------

## 📐 Operators on Vectors

One line of R operates on ALL values at once:

```{r}
systolic_bp <- c(118, 125, 132, 145, 110, 138)
normal_avg  <- 120   # average normal systolic BP

# How much above/below average is each patient?
deviation <- systolic_bp - normal_avg
deviation

# Flag patients above 140 mmHg (hypertension threshold)
systolic_bp > 140
```

> 💡 In Excel, you would write this formula in 6 separate cells. In R, one line handles all patients at once.

------------------------------------------------------------------------

## 🔄 Your Turn — Day 1 Final Exercise

*(5 minutes)*

```{r}
# A clinic measured fasting blood glucose (mg/dL) for 8 patients.
# Patients 1–4: standard diet group
# Patients 5–8: low-carb diet group
# Normal fasting glucose reference: 90 mg/dL

glucose <- c(95, 102, 88, 99, 78, 83, 81, 76)

# 1. How many patients are there?


# 2. What is the average glucose across all patients?


# 3. Calculate how much each patient deviates from the 90 mg/dL reference
deviation

# 4. Average for the standard diet group (patients 1–4)


# 5. Average for the low-carb group (patients 5–8)


# 6. Is the low-carb group average lower?


# 7. Create a character vector for diet_groups
c("Standard","Standard","Standard","Standard",
                "Low-Carb","Low-Carb","Low-Carb","Low-Carb")


```

------------------------------------------------------------------------

## 🏠 Homework Before Day 2

1.  Create a vector of 5 **body temperature** readings (°C) from a pretend patient group
2.  Calculate the `mean`, `max`, and `min`
3.  Create a character vector with patient IDs: `"P01"`, `"P02"`, `"P03"`, `"P04"`, `"P05"`
4.  Write a comment at the top of your script describing the patient group

------------------------------------------------------------------------

## 📌 Day 1 Recap

| Concept           | What You Learned                                        |
|-------------------|---------------------------------------------------------|
| R vs RStudio      | R is the language; RStudio is the workspace             |
| Script vs Console | Write in Script, output appears in Console              |
| Variables         | Store values with `<-`                                  |
| Data Types        | numeric, character, logical, integer, factor            |
| Vectors           | Store multiple values with `c()`                        |
| Operators         | Arithmetic, relational (`>`, `==`), logical (`&`, `\|`) |
