---
title: "R in Action: Code, Analyze & Contribute"
subtitle: "Day 2 — Working with Data Frames | Participant Workbook"
author: "R User Group Nepal × KU Biotechnology Creatives"
date: "April 10, 2026"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

# DAY 2 — Friday, April 10

**R User Group Nepal × Kathmandu University Biotechnology Creatives** *Online · 8:00 PM – 10:00 PM (Nepal Standard Time)*

**What you will be able to do after today:**

-   Write and use functions — including writing a simple one yourself
-   Create and explore data frames — the main data structure in R
-   Extract specific rows and columns
-   Add new calculated columns
-   Add rows and merge two data frames together

------------------------------------------------------------------------

## 🔁 Quick Recap — Day 1

Answer these before we start:

1.  What is the difference between R and RStudio?
2.  What does `<-` do?
3.  What is the difference between `=` and `==`?
4.  What does `class()` tell you?

------------------------------------------------------------------------

## 🔧 Functions

You have already been using functions: `mean()`, `length()`, `class()` are all functions.

A **function** is a named, reusable block of code that takes an input and produces an output.

```{r}
# Functions you already know:
mean(c(118, 142, 112))    # average BP
length(c("A", "B", "C")) # count
class(TRUE)               # data type
```

```{r}
# More useful functions:
round(98.612, digits = 1) # round to 1 decimal → 98.6
sqrt(225)                  # square root → 15
abs(-12.5)                 # absolute value → 12.5
nchar("Kathmandu")        # count characters → 9
toupper("anita")           # uppercase → "ANITA"
```

**Writing your own function:**

When you repeat the same calculation many times, make it a function — write once, use everywhere.

```{r}
# BMI = weight(kg) / height(m)^2
bmi_calculator <- function(weight, height) {
  bmi <- weight / height ^ 2
  return(bmi)
}

# Use it for any patient:
bmi_calculator(60, 1.65)    # → 22.0
bmi_calculator(80, 1.75)    # → 26.1
```

------------------------------------------------------------------------

## 🔄 Your Turn 01 — Functions

*(3 minutes)*

```{r}
# 1. Round 36.746 to 1 decimal place (body temperature)
round( , digits = )

# 2. What is the square root of 400?
sqrt()

# 3. How many characters in "Bhaktapur"?
nchar()

```

```{r}

# 4. Write a function called pulse_pressure that calculates:
#    pulse_pressure = systolic - diastolic
#    Test it: systolic = 118, diastolic = 78

pulse_pressure <- function(systolic, diastolic) {
  
  

}



pulse_pressure(118, 78)    # expected: 40
pulse_pressure(158, 95)    # expected: 63

```

```{r}

# 5. Write a function called fahrenheit_to_celsius
#    Formula: C = (F - 32) * 5/9
#    Test it with 98.6°F (normal body temperature → should give 37°C)







```

------------------------------------------------------------------------

## 📋 Today's Dataset — Patient Vital Signs

> **Dataset description:** 5 patients visited a community health clinic for a routine checkup. We recorded their name, age, systolic blood pressure (mmHg), resting heart rate (bpm), fasting blood glucose (mg/dL), and whether they have been diagnosed with diabetes.
>
> -   **Systolic BP (mmHg):** Normal \< 120 · Elevated 120–129 · High ≥ 130 · Hypertensive crisis \> 180
> -   **Glucose (mg/dL):** Normal \< 100 · Pre-diabetic 100–125 · Diabetic ≥ 126

| patient | age | systolic | heart_rate | glucose | is_diabetic |
|---------|-----|----------|------------|---------|-------------|
| Anita   | 34  | 118      | 72         | 95      | FALSE       |
| Bikash  | 52  | 142      | 88         | 148     | TRUE        |
| Chitra  | 28  | 112      | 68         | 88      | FALSE       |
| Dev     | 61  | 158      | 92         | 172     | TRUE        |
| Elina   | 45  | 135      | 80         | 115     | FALSE       |

> We will keep building on this dataset throughout the day.

------------------------------------------------------------------------

## 📋 Data Frames — Your Main Tool

A **data frame** is a table. Rows are patients (observations). Columns are measurements (variables).

Think of it as an Excel spreadsheet — but controlled entirely by code.

```{r}
# Create this dataset — type it exactly as above
data.frame()

```


```{r}
vital_signs              # please run above dataframe

# These functions help you understand any new dataset
head(vital_signs, 2)        # first 2 rows

```

```{r}
tail(vital_signs, 2)        # last 2 rows
```

```{r}
nrow(vital_signs)        # number of rows (how many patients?)
```

```{r}
ncol(vital_signs)        # number of columns (how many variables?)
dim(vital_signs)         # rows AND columns
names(vital_signs)       # column names
```

```{r}
str(vital_signs)         # structure: data type of each column
```

```{r}
summary(vital_signs)     # basic statistics for every column
```

> 🏥 **Real habit:** Every time you load a new dataset — always run `head()`, `str()`, and `summary()` first. These three tell you almost everything before you start analysis.

------------------------------------------------------------------------

## 🔍 Extracting Data from a Data Frame

**Method 1 — The `$` operator** (to get one full column):

```{r}
vital_signs$patient      # all patient names
vital_signs$systolic     # all systolic BP values
vital_signs$glucose      # all glucose values
```

**Method 2 — Square brackets `[row, column]`** (specific rows or cells):

```{r}
vital_signs[1, ]         # entire first row (Anita's record)
vital_signs[, 3]         # entire third column (systolic)
vital_signs[2, 3]        # row 2, column 3 — one value
vital_signs[1:3, ]       # rows 1 through 3

# Filter rows by condition
vital_signs[vital_signs$systolic >= 140, ]        # high BP patients
vital_signs[vital_signs$is_diabetic == TRUE, ]    # diabetic patients
vital_signs[vital_signs$age < 40, ]               # younger than 40
```

------------------------------------------------------------------------

## 🔄 Your Turn 02 — Exploring Data Frames

*(3 minutes)*

```{r}
# Use vital_signs for all of these.

# 1. Print the whole dataset — how many rows and columns?
vital_signs

# 2. Show only the patient names column
vital_signs$

# 3. Show the entire second row (Bikash's data)
vital_signs[ , ]

# 4. Show all columns for patients with systolic > 130
vital_signs[vital_signs$ > 130, ]

# 5. What is the average systolic BP?
mean(vital_signs$)

# 6. What is the highest glucose reading?
max(vital_signs$)
```

------------------------------------------------------------------------

## ➕ Add a New Column

Adding a calculated column is one of the most common operations in health data analysis.

```{r}
# Classify BP: "High" if systolic >= 140, "Normal" otherwise
vital_signs$bp_status <- ifelse(
  vital_signs$systolic >= 140,
  "High",
  "Normal"
)

vital_signs

# Classify glucose: "High" if > 126 mg/dL (diabetic threshold)
vital_signs$glucose_status <- ifelse(
  vital_signs$glucose > 126,
  "High",
  "Normal"
)

# Flag patients with BOTH high BP and high glucose (dual risk)
vital_signs$dual_risk <- vital_signs$systolic >= 140 &
                         vital_signs$glucose > 126

vital_signs
```

> 💡 `ifelse(CONDITION, value_if_TRUE, value_if_FALSE)` — a compact way to label rows based on a condition. You will use this constantly.

------------------------------------------------------------------------

## 🔄 Your Turn 03 — Add Columns

*(3 minutes)*

```{r}
# Using vital_signs:

# 1. Add a column "high_hr": TRUE if heart_rate > 85
vital_signs$high_hr <- vital_signs$ > 85

# 2. Add a column "age_group":
#    "Senior" if age >= 50, "Adult" otherwise
vital_signs$age_group <- ifelse(
  vital_signs$ >= 50,
  "",
  ""
)

# 3. View the updated data frame — how many columns now?
vital_signs
```

------------------------------------------------------------------------

## ➕ Add a New Row

When a new patient record arrives:

```{r}
# New patient just registered
new_patient <- data.frame(
  patient     = "Farhan",
  age         = 38,
  systolic    = 128,
  heart_rate  = 76,
  glucose     = 102,
  is_diabetic = FALSE
)

# Stack the new row onto the existing table
vital_signs_v2 <- rbind(vital_signs, new_patient)

nrow(vital_signs_v2)    # should now be 6
tail(vital_signs_v2)    # check the last row
```

> ⚠️ The new row must have the **exact same column names and types** as the original. Use the base `vital_signs` (before you added bp_status etc.) for `rbind()`.

------------------------------------------------------------------------

## 🔗 Merge Two Data Frames

In real health work, data often lives in separate tables. You join them by a shared column.

```{r}
# Table 1: Patient registry
patients_info <- data.frame(
  patient_id = c("P01","P02","P03"),
  name       = c("Anita","Bikash","Chitra"),
  district   = c("Kathmandu","Lalitpur","Bhaktapur")
)

# Table 2: Blood test results (from a separate lab system)
blood_tests <- data.frame(
  patient_id  = c("P01","P02","P03"),
  hemoglobin  = c(12.1, 14.5, 11.8),
  cholesterol = c(185, 240, 172)
)

# Join them by the "patient_id" column
merged_data <- merge(patients_info, blood_tests, by = "patient_id")
merged_data
```

> **Result:** Only patients present in **both** tables appear (inner join by default). You will learn all join types on Day 3.

------------------------------------------------------------------------

## ✏️ Rename a Column

```{r}
# Check current names
names(vital_signs)

# Rename "systolic" to "systolic_bp"
names(vital_signs)[names(vital_signs) == "systolic"] <- "systolic_bp"

# Or rename by position (column 3)
names(vital_signs)[3] <- "systolic_bp"
```

------------------------------------------------------------------------

## 🔄 Your Turn 04 — Build a Full Dataset

*(5 minutes)*

> **Context:** A health post collected data from 4 patients on two different diets. Normal fasting glucose = 90 mg/dL.

```{r}

# Step 1: Create the data frame
diet_study <- data.frame(
  patient    = c("Gita","Hari","Isha","Jay"),
  diet_type  = c("Standard","Standard","Low-Carb","Low-Carb"),
  glucose    = c(105, 98, 82, 79),
  systolic   = c(128, 135, 118, 122)
)

# Step 2: View it — how many rows? how many columns?

# Step 3: Add a column "glucose_diff" = glucose - 90

# Step 4: Add a column "above_normal" = TRUE if glucose > 90

# Step 5: Show only the "Low-Carb" rows

# Step 6: What is the average glucose for Low-Carb patients?

# Step 7: Rename "systolic" to "systolic_bp"
```

------------------------------------------------------------------------

## ☕ Break — 5 Minutes

------------------------------------------------------------------------

## 🔁 Day 2 Recap & What's Next

**Today you learned:**

| Skill | How You Used It |
|------------------------------------|------------------------------------|
| Functions | Used `mean()`, `round()`, wrote `bmi_calculator()` and `pulse_pressure()` |
| Data Frames | Created `vital_signs`, explored with `head()`, `str()`, `summary()` |
| Extract Data | Used `$`, `[ , ]`, and conditions to filter rows |
| Add Columns | Used `$` assignment and `ifelse()` |
| Add Rows | Used `rbind()` |
| Merge Tables | Used `merge()` to join two data frames |
| Rename Columns | Used `names()` |

**Day 3 (In-person, Saturday):**

-   Load real data from CSV and Excel files
-   Write cleaner code using the pipe `%>%`
-   Use `dplyr` for professional data manipulation
-   Reshape messy data with `tidyr`

------------------------------------------------------------------------

## 📎 Quick Reference — Day 1 & Day 2

```{r}
#| eval: false
# ── VARIABLES ──────────────────────────────
x            <- 42
patient_name <- "Anita"
is_diabetic  <- FALSE

# ── VECTORS ────────────────────────────────
systolic <- c(118, 125, 132, 145)
length(systolic)         # count
mean(systolic)           # average
max(systolic)            # highest
systolic[1]              # first element
systolic[1:3]            # first three elements
systolic - 120           # subtract from all at once

# ── DATA TYPES ─────────────────────────────
class(118.5)    # "numeric"
class("Anita")  # "character"
class(TRUE)     # "logical"
class(50L)      # "integer"

# ── DATA FRAME ─────────────────────────────
df <- data.frame(name = c("A","B"), bp = c(118, 128))
head(df)                         # first rows
str(df)                          # structure
summary(df)                      # quick stats
nrow(df); ncol(df); names(df)    # dimensions

df$name                          # one column
df[1, ]                          # one row
df[df$bp > 120, ]                # filtered rows
df$new_col <- df$bp - 110        # add column
rbind(df, new_row)               # add row

# ── MERGE ──────────────────────────────────
merge(df1, df2, by = "patient_id")

# ── RENAME ─────────────────────────────────
names(df)[names(df) == "bp"] <- "systolic_bp"

# ── FUNCTIONS ──────────────────────────────
my_func <- function(x, y) { return(x - y) }
my_func(118, 78)    # → 40

# ── IFELSE ─────────────────────────────────
ifelse(df$bp >= 140, "High", "Normal")
```
