---
title: "R in Action: Code, Analyze & Contribute"
subtitle: "Day 4 — RNA-seq Case Study | Participant Workbook"
author: "R User Group Nepal × KU Biotechnology Creatives"
date: "April 12, 2026"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

------------------------------------------------------------------------

## 🧬 RNA-seq Case Study

------------------------------------------------------------------------

> *"Everything you built today with health data — layers, facets, line plots, patchwork — is exactly the toolkit bioinformaticians use with genomics data."*

This case study uses RNA-seq data from **Fu et al. 2015** (GEO: GSE60450). The study examined gene expression in **basal** and **luminal** mammary cells from mice at three stages: virgin, pregnant, and lactating. There are **2 samples per group × 6 groups = 12 samples total**.

### Key RNA-seq Terminology

| Term | Meaning | Health Analogy |
|----|----|----|
| Gene | Unit of DNA coding for a protein | A specific blood test |
| Expression | How actively a gene is used | The test result value |
| Count / CPM | Normalized RNA fragments detected | The measurement reading |
| Sample | One biological specimen | One patient visit |
| Condition | Experimental group | Control vs Treatment |
| Replicate | Repeated measurement | Duplicate readings |
| Differential expression | Gene significantly up/down | Abnormal lab value |
| log2 fold change | Magnitude of change | 2× higher / lower |
| p-value / FDR | Statistical confidence | Same as in any test |

| Group Code | Cell Type | Stage     |
|------------|-----------|-----------|
| `bvirg`    | Basal     | Virgin    |
| `bpreg`    | Basal     | Pregnant  |
| `blact`    | Basal     | Lactating |
| `lvirg`    | Luminal   | Virgin    |
| `lpreg`    | Luminal   | Pregnant  |
| `llact`    | Luminal   | Lactating |

------------------------------------------------------------------------

```{r}
library(tidyverse)
```

------------------------------------------------------------------------

## PART 1 — Load and Explore

------------------------------------------------------------------------

### Explore functions reference

| Function      | What it shows                       |
|---------------|-------------------------------------|
| `head(df)`    | First 6 rows                        |
| `tail(df)`    | Last 6 rows                         |
| `dim(df)`     | Number of rows and columns          |
| `names(df)`   | Column names                        |
| `str(df)`     | Column types and first values       |
| `summary(df)` | Min, Max, mean, NAs per column      |
| `df$col[1:3]` | First 3 values of a specific column |

------------------------------------------------------------------------

### 🔄 Your Turn CS-01 — Load & Explore *(5 minutes)*

> **Scenario:** Two CSV files are waiting in the `data/` folder. Load them, rename the first column of each, and do a quick exploration to understand what you have.

```{r}
# Load the counts file (CPM and TMM normalized data)
# Filename: GSE60450_GeneLevel_Normalized(CPM.and.TMM)_data.csv
counts <- read_csv("data/GSE60450_GeneLevel_Normalized(CPM.and.TMM)_data.csv",
                   show_col_types = FALSE)
```

```{r}
# Load the sample metadata file
# Filename: GSE60450_filtered_metadata.csv
sampleinfo <- read_csv("data/GSE60450_filtered_metadata.csv",
                       show_col_types = FALSE)
```

```{r}
# Explore sampleinfo — run at least 3 of: head(), dim(), tail(), names(), str()
head(sampleinfo)
dim(sampleinfo)
str(sampleinfo)

# Look at the first 3 values of the characteristics column
# This tells us what each sample is
sampleinfo$characteristics[1:3]
```

```{r}
# Rename first column of sampleinfo → "sample_id"
# Rename first column of counts    → "gene_id"
colnames(sampleinfo)[1] <- "sample_id"
colnames(counts)[1]     <- "gene_id"
```

```{r}
# Check the expression range across samples
summary(counts)
# The Min and Max values tell you why we need a log2 transform
```

------------------------------------------------------------------------

## PART 2 — Reshape Wide to Long

------------------------------------------------------------------------

The counts file has **one column per sample** (wide format). ggplot2 needs **one row per gene per sample** (long format). There are three ways to do this with `pivot_longer()`.

```{r}
# Method 1: select columns whose names start with "GSM"
seqdata <- pivot_longer(
  counts,
  cols      = starts_with("GSM"),
  names_to  = "Sample",
  values_to = "Count"
)

seqdata
nrow(seqdata)    # should be n_genes × 12 samples
```

```{r}
# Method 2: select columns by range (first to last GSM column)
seqdata <- pivot_longer(
  counts,
  cols      = GSM1480291:GSM1480302,
  names_to  = "Sample",
  values_to = "Count"
)

seqdata
```

```{r}
# Method 3: exclude the two identifier columns (RECOMMENDED)
# This is the most robust approach — works even if sample names change
seqdata <- pivot_longer(
  counts,
  cols      = -c("gene_id", "gene_symbol"),
  names_to  = "Sample",
  values_to = "Count"
)

nrow(seqdata)    # n_genes × 12 samples
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-02 — Reshape *(5 minutes)*

> **Scenario:** Practice all three `pivot_longer()` methods and understand why Method 3 is recommended.

```{r}
# Method 1: cols = starts_with("GSM")
# Run it and check nrow() of the result
seqdata_m1 <- pivot_longer(
  counts,
  cols      = starts_with("GSM"),
  names_to  = "Sample",
  values_to = "Count"
)
nrow(seqdata_m1)
```

```{r}
# Method 2: cols = GSM1480291:GSM1480302
seqdata_m2 <- pivot_longer(
  counts,
  cols      = GSM1480291:GSM1480302,
  names_to  = "Sample",
  values_to = "Count"
)
nrow(seqdata_m2)
```

```{r}
# Method 3 (recommended): cols = -c("gene_id", "gene_symbol")
seqdata <- pivot_longer(
  counts,
  cols      = -c("gene_id", "gene_symbol"),
  names_to  = "Sample",
  values_to = "Count"
)
nrow(seqdata)
```

```{r}
# Task ★ — Explain
# Do all three methods give the same nrow()?
# Why is Method 3 the most robust for real-world data?
# When would Method 1 fail?

# Your explanation here:
# Method 3 is robust because...
```

------------------------------------------------------------------------

## PART 3 — Join and Quality Check

------------------------------------------------------------------------

### Why log2(Count + 1)?

| Scale | Problem |
|----|----|
| Raw `Count` | Extreme values (0 to 50,000+) squash low-expression genes to the bottom |
| `log2(Count)` | `log2(0) = -Inf` — produces warnings for genes with zero reads |
| `log2(Count + 1)` | The `+1` ensures zero counts become `log2(1) = 0` — the standard approach |

------------------------------------------------------------------------

### 🔄 Your Turn CS-03 — Join & QC Plot *(5 minutes)*

> **Scenario:** Combine expression data with sample metadata using `full_join()`, then make a quality-check boxplot to confirm all 12 samples are comparable.

```{r}
# Join seqdata (Sample column) with sampleinfo (sample_id column)
# full_join keeps all rows from both tables
allinfo <- full_join(
  seqdata,
  sampleinfo,
  by = join_by(Sample == sample_id)
)

allinfo
```

```{r}
# QC boxplot — linear scale first to see the problem
ggplot(data = allinfo,
       mapping = aes(x = Sample, y = Count)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(title = "Linear scale — extreme values dominate")
# What do you notice about the y-axis?
```

```{r}
# log2 scale — but log2(0) produces warnings
ggplot(data = allinfo,
       mapping = aes(x = Sample, y = log2(Count))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
# Any warnings in the console?
```

```{r}
# log2(Count + 1) — the +1 prevents -Inf for zero counts (standard approach)
ggplot(data = allinfo,
       mapping = aes(x = Sample, y = log2(Count + 1))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(title = "Quality check: all samples comparable?",
       y     = "log2(CPM + 1)")
```

```{r}
# Add fill = Sample to colour each box differently
# Does this make it easier to spot any outlier sample?
ggplot(data = allinfo,
       mapping = aes(x = Sample, y = log2(Count + 1), fill = Sample)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

```{r}
# Task ★ — Interpret
# Are all 12 boxes at a similar median height?
# What would it mean biologically if one sample's box was much higher or lower?

# Your interpretation:
```

------------------------------------------------------------------------

## PART 4 — Label the Groups

------------------------------------------------------------------------

The `characteristics` column contains long strings like `"cell type: basal; developmental stage: virgin"`. We need short group codes.

```{r}
# str_detect() checks if a string contains a pattern
# "basal.*virgin" means: "basal" followed by anything (.*), then "virgin"
str_detect("cell type: basal; developmental stage: virgin",
           "basal.*virgin")   # → TRUE
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-04 — Label Groups *(5 minutes)*

> **Scenario:** Create short group labels using `mutate()` + `case_when()` + `str_detect()`.

```{r}
# Complete the case_when to create a Group column with 6 short codes
# The pattern "basal.*preg" means: "basal" followed anywhere by "preg"
allinfo <- mutate(allinfo, Group = case_when(
  str_detect(characteristics, "basal.*virgin")   ~ "bvirg",
  str_detect(characteristics, "basal.*preg")     ~ "bpreg",
  str_detect(characteristics, "basal.*lact")     ~ "blact",
  str_detect(characteristics, "luminal.*virgin") ~ "lvirg",
  str_detect(characteristics, "luminal.*preg")   ~ "lpreg",
  str_detect(characteristics, "luminal.*lact")   ~ "llact"
))
```

```{r}
# Check: how many rows per group?
# Expected: 2 samples × n_genes rows per group
table(allinfo$Group)
```

```{r}
# Task ★ — Pattern matching
# Does "basal.*pregnant" also work? Try replacing "basal.*preg" and check table()
# What does this tell you about how str_detect() matches patterns?

```

------------------------------------------------------------------------

## PART 5 — Select Genes of Interest

------------------------------------------------------------------------

Two approaches: specify genes you already know are interesting (Method A), or let the data identify the most expressed genes (Method B).

### Method A — Manual list

```{r}
# Biologically motivated list — these are known milk protein and housekeeping genes
mygenes <- c("Csn1s2a", "Csn1s1", "Csn2", "Glycam1",
             "COX1", "Trf", "Wap", "Eef1a1")

mygenes_counts <- filter(allinfo, gene_symbol %in% mygenes)
mygenes_counts
```

### Method B — Top-n by total expression

```{r}
# Data-driven: group → summarise → arrange → head → pull
mygenes <- allinfo %>%
  group_by(gene_symbol) %>%
  summarise(total_count = sum(Count)) %>%
  arrange(desc(total_count)) %>%
  head(n = 8) %>%
  pull(gene_symbol)

mygenes    # see which genes are most expressed across all samples

mygenes_counts <- filter(allinfo, gene_symbol %in% mygenes)
mygenes_counts
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-05 — Select Genes *(5 minutes)*

> **Scenario:** Compare both approaches and understand when to use each.

```{r}
# Method A: filter to the manual list
mygenes_manual <- c("Csn1s2a","Csn1s1","Csn2","Glycam1",
                    "COX1","Trf","Wap","Eef1a1")

mygenes_counts_a <- filter(allinfo, gene_symbol %in% mygenes_manual)
nrow(mygenes_counts_a)    # how many rows? (8 genes × 12 samples)
```

```{r}
# Method B: top 8 most expressed genes
mygenes_top <- allinfo %>%
  group_by(gene_symbol) %>%
  summarise(total_count = sum(Count)) %>%
  arrange(desc(total_count)) %>%
  head(n = 8) %>%
  pull(gene_symbol)

mygenes_top
mygenes_counts <- filter(allinfo, gene_symbol %in% mygenes_top)
```

```{r}
# Task ★ — Compare: which genes are in one list but not the other?
setdiff(mygenes_manual, mygenes_top)   # in A, not in B
setdiff(mygenes_top, mygenes_manual)   # in B, not in A

# What does this tell you about the two approaches?
# Your interpretation:
```

------------------------------------------------------------------------

## PART 6 — Build the Expression Plot

------------------------------------------------------------------------

Build the plot incrementally — add one layer at a time and observe how the visualization improves.

```{r}
# Step 1: Boxplot — a familiar starting point
ggplot(data = mygenes_counts,
       mapping = aes(x = Group, y = log2(Count + 1), fill = Group)) +
  geom_boxplot() +
  facet_wrap(~ gene_symbol)
```

```{r}
# Step 2: Switch to geom_point() — individual data points
ggplot(data = mygenes_counts,
       mapping = aes(x = Group, y = log2(Count + 1))) +
  geom_point() +
  facet_wrap(~ gene_symbol)
# Problem: points overlap — can you tell how many points are at each position?
```

```{r}
# Step 3: Switch to geom_jitter() — adds random horizontal offset to separate points
ggplot(data = mygenes_counts,
       mapping = aes(x = Group, y = log2(Count + 1))) +
  geom_jitter() +
  facet_wrap(~ gene_symbol)
```

```{r}
# Step 4: Add colour = Group inside aes()
ggplot(data = mygenes_counts,
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol)
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-06 — Build the Plot *(8 minutes)*

> **Scenario:** Build the complete multi-gene expression figure step by step.

```{r}
# Task 1: Boxplot with fill = Group and facet_wrap(~ gene_symbol)

```

```{r}
# Task 2: Replace geom_boxplot with geom_point, then geom_jitter
# What changes each time?

```

```{r}
# Task 3: Add colour = Group to aes(), set width = 0.1 in geom_jitter()

```

```{r}
# Task 4: Apply custom colours
mycolours <- c("turquoise","plum","tomato","violet","steelblue","chocolate")

# Add scale_colour_manual(values = mycolours)

```

```{r}
# Task 5 ★ — Combine layers
# Add geom_boxplot(alpha = 0.2) BEFORE geom_jitter()
# Which layer goes first? Why does the order matter?
# (Hint: layers render in the order they appear — first = bottom, last = top)

```

------------------------------------------------------------------------

## PART 7 — Themes, Labels, and Saving

------------------------------------------------------------------------

### Theme options for RNA-seq plots

```{r}
# Base plot to apply themes to
p <- ggplot(mygenes_counts,
            aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = mycolours) +
  labs(x = "Cell type and stage", y = "log2(Count + 1)",
       title = "Mammary gland RNA-seq data")
```

```{r}
# theme_bw() — white background with gridlines
p + theme_bw() + theme(axis.text.x = element_text(angle = 90))
```

```{r}
# theme_minimal() — very clean, light gridlines
p + theme_minimal() + theme(axis.text.x = element_text(angle = 90))
```

```{r}
# Manual theme() — remove background and grid explicitly
# Equivalent to theme_classic() but shows you what's happening under the hood
p + theme(
  axis.text.x      = element_text(angle = 90),
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
)
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-07 — Themes & Saving *(5 minutes)*

> **Scenario:** Choose a theme, add proper labels, rotate x-axis text, and save in two formats.

```{r}
# Task 1: Compare themes
# Run the base plot p with theme_bw(), then theme_minimal(), then manual theme()
# Which do you prefer for a publication?

```

```{r}
# Task 2: Add axis rotation
# Add theme(axis.text.x = element_text(angle = 90)) to your preferred theme
# Why is rotation needed here specifically?

```

```{r}
# Task 3 ★ — Save as PNG using ggsave()
# ggsave() saves the LAST plot that was printed
ggsave("plots/mammary_expression.png", width = 10, height = 8, dpi = 300)
```

```{r}
#| eval: false

# Task 4 ★ — Save as PDF using pdf() ... dev.off()
# This method: open the PDF device, print the plot, close the device
pdf("plots/mammary_expression.pdf", width = 10, height = 8)
print(p + theme_classic() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5)))
dev.off()

# Open both files: which format is better for a journal submission?
```

------------------------------------------------------------------------

## PART 8 — Order Groups with `factor()`

------------------------------------------------------------------------

Without factor levels, R sorts groups alphabetically: `blact → bpreg → bvirg → llact → lpreg → lvirg`. This mixes cell types and stages — biologically meaningless.

```{r}
# Define the biologically correct order:
# virgin → pregnant → lactating, for each cell type
group_order <- c("bvirg", "bpreg", "blact", "lvirg", "lpreg", "llact")

mygenes_counts <- mutate(
  mygenes_counts,
  Group_f = factor(Group, levels = group_order)
)
```

```{r}
# Verify the levels are set correctly
levels(mygenes_counts$Group_f)
```

```{r}
# Inspect: compare original character column with the new factor column
select(mygenes_counts, gene_id, Group, Group_f)
```

```{r}
# str() confirms Group_f is stored as a Factor with 6 ordered levels
str(mygenes_counts)
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-08 — Final Publication Plot *(8 minutes)*

> **Scenario:** Build the complete publication-quality figure using `Group_f` on the x-axis.

```{r}
# Task 1: Build the final plot using Group_f on x-axis
# Include all elements:
# - geom_jitter(size = 3, width = 0.15)
# - facet_wrap(~ gene_symbol)
# - scale_colour_manual(values = mycolours)
# - labs() with title, subtitle, x, y labels
# - theme_classic()
# - theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
# - theme(strip.text = element_text(face = "bold"))

ggplot(mygenes_counts,
       aes(x = Group_f, y = log2(Count + 1), colour = Group_f)) +
  geom_jitter(size = 3, width = 0.15) +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = mycolours) +
  labs(x        = "Cell Type and Stage",
       y        = "Expression (log2 CPM + 1)",
       title    = "Mammary Gland Gene Expression",
       subtitle = "Fu et al. 2015, GSE60450") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text  = element_text(face = "bold"))
```

```{r}
#| eval: false
# Task 2: Save as PNG
ggsave("plots/mammary_expression.png", width = 10, height = 8, dpi = 300)
```

```{r}
#| eval: false
# Task 3: Save as PDF
pdf("plots/mammary_expression.pdf", width = 10, height = 8)
ggplot(mygenes_counts,
       aes(x = Group_f, y = log2(Count + 1), colour = Group_f)) +
  geom_jitter(size = 3, width = 0.15) +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = mycolours) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
dev.off()
```

```{r}
# Task 4 ★ — Biological interpretation
# Look at the final plot and answer:
# 1. Which genes are highest in the lactating stage (blact / llact)?
# 2. Which gene shows the most consistent expression across all groups?
# 3. What is the biggest expression difference between basal and luminal cells?

# Your interpretation:
```

------------------------------------------------------------------------

### ✅ Exercises

**Exercise 1:** Replace `mycolours` with 6 colors of your choice. Try `scale_colour_brewer(palette = "Set2")` as an alternative — does it work? Why not?

**Exercise 2:** Change `facet_wrap(~ gene_symbol)` to `ncol = 2`. What changes in the layout?

**Exercise 3:** Filter `mygenes_counts` to only luminal groups (`lvirg`, `lpreg`, `llact`) and remake the full plot.

**Exercise 4:** Add `geom_hline(yintercept = 8, linetype = "dashed", colour = "gray40")` to the plot. What does `log2(Count + 1) = 8` correspond to in CPM? *(Hint: 2\^8 − 1 = ?)*

------------------------------------------------------------------------

## 🌋 Volcano Plot — Differential Expression

------------------------------------------------------------------------

### What Is a Volcano Plot?

A volcano plot shows **two statistics simultaneously** for every gene:

-   **X-axis:** `log2 fold change` — direction and size of change (Basal vs Luminal)
-   **Y-axis:** `-log10(adjusted p-value)` — statistical confidence

Genes that are both **large change** (far left or right) AND **statistically significant** (high up) appear in the "eruption" — these are the biologically interesting genes.

| Region                | Meaning                                    |
|-----------------------|--------------------------------------------|
| Top-right             | Significantly **up-regulated** in Basal    |
| Top-left              | Significantly **up-regulated** in Luminal  |
| Bottom-middle         | Changed but not significant — likely noise |
| Far left/right, low y | Large change but not reliable              |

### Why `-log10(p-value)`?

```{r}
# p-values become very small — hard to read on a linear scale
# -log10 flips them: larger number = more significant
-log10(c(0.05, 0.01, 0.001, 0.0001))
# →     1.3    2.0   3.0    4.0
# Significance threshold: -log10(0.05) ≈ 1.3
```

------------------------------------------------------------------------

### 🔄 Your Turn CS-09 — Volcano Plot *(8 minutes)*

> **Scenario:** Build the signature DE figure step by step.

#### Step 1 — Load DE Results

```{r}
de_results <- read_csv("data/de_results.csv", show_col_types = FALSE)
head(de_results)
glimpse(de_results)
```

> Columns: `gene_name`, `log2_fc` (Basal vs Luminal), `p_value`, `p_adjusted` (FDR-corrected)

#### Step 2 — Add Significance Category

```{r}
# Add neg_log10_p and significance using mutate() + case_when()
# Cutoffs: p_adjusted < 0.05 AND abs(log2_fc) > 1
de_results <- de_results %>%
  mutate(
    neg_log10_p  = -log10(p_adjusted),
    significance = case_when(
      p_adjusted < 0.05 & log2_fc >  1 ~ "Up in Basal",
      p_adjusted < 0.05 & log2_fc < -1 ~ "Up in Luminal",
      TRUE                              ~ "Not significant"
    )
  )

# How many genes fall into each category?
de_results %>% count(significance)
```

#### Step 3 — Basic Volcano

```{r}
# Scatter: x = log2_fc, y = neg_log10_p, colour = significance
ggplot(de_results,
       aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "steelblue",
               "Not significant" = "gray70")
  ) +
  theme_classic()
```

#### Step 4 — Add Threshold Lines

```{r}
# Add dashed lines: horizontal at -log10(0.05), vertical at -1 and +1
ggplot(de_results,
       aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             colour = "gray40") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             colour = "gray40") +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "steelblue",
               "Not significant" = "gray70")
  ) +
  theme_classic()
```

#### Step 5 ★ — Label Top Genes & Save

```{r}
# Select the 10 most significant genes (highest neg_log10_p) among significant ones
top_sig <- de_results %>%
  filter(significance != "Not significant") %>%
  slice_max(neg_log10_p, n = 10)

# Build the final labeled volcano
ggplot(de_results,
       aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             colour = "gray40") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             colour = "gray40") +
  geom_text(data = top_sig, aes(label = gene_name),
            size = 2.8, nudge_y = 0.25, colour = "black") +
  scale_colour_manual(
    values = c("Up in Basal"     = "tomato",
               "Up in Luminal"   = "steelblue",
               "Not significant" = "gray70")
  ) +
  labs(x        = "log2 Fold Change (Basal vs Luminal)",
       y        = "-log10(adjusted p-value)",
       title    = "Differential Expression: Basal vs Luminal Cells",
       subtitle = "Fu et al. 2015 · |log2FC| > 1, FDR < 0.05",
       colour   = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

```{r}
#| eval: false
ggsave("plots/volcano_basal_vs_luminal.png", width = 8, height = 6, dpi = 300)
```

------------------------------------------------------------------------

## 📎 Full Case Study Quick Reference

```{r}
#| eval: false

# ── LOAD ────────────────────────────────────────────────────────────────────
counts     <- read_csv("data/GSE60450_GeneLevel_Normalized(CPM.and.TMM)_data.csv")
sampleinfo <- read_csv("data/GSE60450_filtered_metadata.csv")
colnames(sampleinfo)[1] <- "sample_id"
colnames(counts)[1]     <- "gene_id"

# ── RESHAPE — three equivalent methods ──────────────────────────────────────
seqdata <- pivot_longer(counts, cols = starts_with("GSM"),
                        names_to = "Sample", values_to = "Count")
seqdata <- pivot_longer(counts, cols = GSM1480291:GSM1480302,
                        names_to = "Sample", values_to = "Count")
seqdata <- pivot_longer(counts, cols = -c("gene_id","gene_symbol"),
                        names_to = "Sample", values_to = "Count")

# ── JOIN ─────────────────────────────────────────────────────────────────────
allinfo <- full_join(seqdata, sampleinfo,
                     by = join_by(Sample == sample_id))

# ── QC BOXPLOT ───────────────────────────────────────────────────────────────
ggplot(allinfo, aes(x = Sample, y = log2(Count + 1))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# ── GROUP LABELS ─────────────────────────────────────────────────────────────
allinfo <- mutate(allinfo, Group = case_when(
  str_detect(characteristics, "basal.*virgin")   ~ "bvirg",
  str_detect(characteristics, "basal.*lact")     ~ "blact",
  str_detect(characteristics, "luminal.*virgin") ~ "lvirg",
  str_detect(characteristics, "luminal.*lact")   ~ "llact"
  # ... (all 6 patterns)
))

# ── SELECT GENES — two methods ───────────────────────────────────────────────
mygenes_counts <- filter(allinfo,
  gene_symbol %in% c("Csn1s2a","Wap","Eef1a1","..."))   # manual

mygenes <- allinfo %>%                                    # top-n
  group_by(gene_symbol) %>%
  summarise(total_count = sum(Count)) %>%
  arrange(desc(total_count)) %>%
  head(n = 8) %>%
  pull(gene_symbol)

# ── EXPRESSION PLOT ──────────────────────────────────────────────────────────
group_order <- c("bvirg","bpreg","blact","lvirg","lpreg","llact")

mygenes_counts <- mutate(mygenes_counts,
  Group_f = factor(Group, levels = group_order))

ggplot(mygenes_counts, aes(x = Group_f, y = log2(Count + 1), colour = Group_f)) +
  geom_jitter(size = 3, width = 0.15) +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = mycolours) +
  labs(x = "Cell Type and Stage", y = "Expression (log2 CPM + 1)",
       title = "Mammary Gland Gene Expression") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# ── SAVE ─────────────────────────────────────────────────────────────────────
ggsave("plots/mammary_expression.png", width = 10, height = 8, dpi = 300)

pdf("plots/mammary_expression.pdf", width = 10, height = 8)
print(last_plot())
dev.off()

# ── VOLCANO ESSENTIALS ───────────────────────────────────────────────────────
de_results <- de_results %>%
  mutate(neg_log10_p = -log10(p_adjusted),
         significance = case_when(...))

ggplot(de_results, aes(x = log2_fc, y = neg_log10_p, colour = significance)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed") +
  geom_text(data = top_sig, aes(label = gene_name), nudge_y = 0.25) +
  scale_colour_manual(values = c("Up in Basal" = "tomato", ...))
```
