---
title: "R in Action: Code, Analyze & Contribute"
subtitle: "Day 3 — Real Data: Import, Clean & Reshape | Participant Workbook"
author: "Binod Jung Bogati - R User Group Nepal"
date: "April 11, 2026"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

# DAY 3 — Saturday, April 11

**In-Person · 9:00 AM – 4:00 PM (Nepal Standard Time)**

**What you will be able to do after today:**

-   Load data from CSV files into R
-   Handle missing values (`NA`) confidently
-   Write clean, readable code using the pipe `%>%`
-   Filter, select, sort, group, and summarize data with `dplyr`
-   Categorize values with `if_else()` and `case_when()`
-   Join two related datasets
-   Reshape data between wide and long format using `tidyr`

> 💡 **How to use this workbook:**\
> - Each **Your Turn** section has starter code with blanks (`___`) or empty for you to fill in.\
> - Read the hint comments carefully.\
> - Run code as you fill it in — do not wait until everything is complete before running.

------------------------------------------------------------------------

## 📋 Today's Dataset — Community Health Screening

> **Dataset description:** A community health screening across three Kathmandu Valley districts. Eight adult patients visited a local health post. We recorded age, gender, systolic BP (mmHg), resting heart rate (bpm), fasting blood glucose (mg/dL), and district.
>
> **Clinical reference ranges:**
>
> \- **Systolic BP (mmHg):** Normal \<120 · Elevated 120–129 · High Stage 1: 130–139 · High Stage 2: ≥140
>
> \- **Glucose (mg/dL):** Normal \<100 · Pre-diabetic 100–125 · Diabetic ≥126

```{r}
library(tidyverse)

patients <- read_csv("data/patients.csv", show_col_types = FALSE)
```

```{r}
head(patients)
```

```{r}
glimpse(patients)
```

```{r}
summary(patients)
```

------------------------------------------------------------------------

## PART 1 — Data Import & Export

------------------------------------------------------------------------

### 📁 RStudio Projects

```         
r_workshop/
├── data/       ← all CSV and Excel files go here
├── scripts/    ← your R scripts go here
├── plots/      ← saved figures go here
└── r_workshop.Rproj   ← always open this to start
```

**Create:** `File → New Project → New Directory → New Project → Name: r_workshop`

> ✅ Always open the `.Rproj` file to start your session.

------------------------------------------------------------------------

### 📥 Loading Data

```{r}
#| eval: false

library(tidyverse)

# Load a CSV file — read_csv() is faster and cleaner than read.csv()
patients <- read_csv("data/patients.csv", show_col_types = FALSE)

# Load an Excel file
library(readxl)
lab_data <- read_excel("data/lab_results.xlsx")
```

```{r}
#| eval: false

# After loading, ALWAYS run these three checks:
head(patients)        # first 6 rows — does it look right?
glimpse(patients)     # column types — are numbers stored as numbers?
summary(patients)     # quick statistics per column
```

------------------------------------------------------------------------

### 📤 Saving Data

```{r}
#| eval: false

# Save as CSV with a new name — NEVER overwrite the original
write_csv(patients, "data/patients_cleaned.csv")
```

------------------------------------------------------------------------

## PART 2 — Handling Missing Values (`NA`)

------------------------------------------------------------------------

> ⚠️ Any calculation that includes `NA` returns `NA` by default. Always check before analyzing.

```{r}
# Load a messy dataset that has missing values
patients_messy <- read_csv("data/patients_messy.csv", show_col_types = FALSE)
patients_messy
```

```{r}
# Detect where NAs are
is.na(patients_messy$age)         # TRUE for each NA position
sum(is.na(patients_messy))        # total NAs across whole dataset
colSums(is.na(patients_messy))    # NAs counted per column
```

```{r}
# NA breaks calculations — na.rm = TRUE fixes it
mean(patients_messy$age)                   # → NA  (broken!)
mean(patients_messy$age, na.rm = TRUE)     # → works correctly
```

```{r}
# Remove ALL rows that have any missing value
patients_messy %>%
  drop_na()
```

```{r}
# Remove rows only where a SPECIFIC column is missing
patients_messy %>%
  drop_na(systolic)
```

```{r}
# Replace NA with a calculated value
patients_messy %>%
  mutate(
    age      = replace_na(age, mean(age, na.rm = TRUE)),
    systolic = replace_na(systolic, median(systolic, na.rm = TRUE))
  )
```

------------------------------------------------------------------------

### 🔄 Your Turn 01 — Missing Values *(5 minutes)*

> **Instructions:** Fill in the blanks (`___`). Each blank has a hint in the comment above it.

```{r}
# Load the survey dataset
survey <- read_csv("data/survey.csv", show_col_types = FALSE)
survey
```

```{r}
# 1. How many NAs are in each column?
#    Hint: colSums(is.na(DATASET_NAME))
colSums(is.na(___))
```

```{r}
# 2. What is the average age? (use na.rm = TRUE to ignore NAs)
#    Hint: mean(survey$age, na.rm = TRUE)
mean(___, na.rm = ___)
```

```{r}
# 3. Drop ALL rows that have any NA — how many rows remain?
survey %>%
  drop_na()
```

```{r}
# 4. Drop rows only where weight_kg is missing
#    Hint: drop_na(COLUMN_NAME)
survey %>%
  drop_na(___)
```

```{r}
# 5. ★ Replace NA in age with the MEDIAN age
#       Replace NA in smoker with "Unknown"
#    Hint: replace_na(age, median(age, na.rm = TRUE))
survey %>%
  mutate(
    age    = replace_na(age, median(___, na.rm = TRUE)),
    smoker = replace_na(smoker, ___)
  )
```

------------------------------------------------------------------------

## PART 3 — The Pipe and `dplyr` Verbs

------------------------------------------------------------------------

The pipe `%>%` passes the result of one step into the next — read it as **"and then"**.

```{r}
#| eval: false

# Without pipe — nested and hard to read
arrange(filter(select(patients, name, age, systolic), systolic >= 140), age)
```

```{r}
# With pipe — reads like steps from top to bottom
patients %>%
  select(name, age, systolic) %>%
  filter(systolic >= 140) %>%
  arrange(age)
```

> ⌨️ Shortcut: **Ctrl + Shift + M** (Windows) / **Cmd + Shift + M** (Mac)

------------------------------------------------------------------------

### `select()` — Choose Columns

```{r}
# Keep specific columns by name
patients %>% select(name, age, systolic)
```

```{r}
# Remove a column with minus sign (-)
patients %>% select(-patient_id)
```

```{r}
# Keep columns whose names match a pattern
patients %>% select(starts_with("g"))    # → glucose, gender
patients %>% select(ends_with("e"))      # → name, glucose, age
```

------------------------------------------------------------------------

### `filter()` — Choose Rows

```{r}
# Hypertensive patients (Stage 2: systolic >= 140)
patients %>% filter(systolic >= 140)
```

```{r}
# Female patients in Kathmandu
patients %>% filter(gender == "F" & district == "Kathmandu")
```

```{r}
# From Lalitpur OR Bhaktapur (%in% checks a list)
patients %>% filter(district %in% c("Lalitpur", "Bhaktapur"))
```

**Logical operators:**

| Operator | Meaning | Example |
|-----------------|------------------------------------------|-------------|
| `==` | Equal to | `gender == "F"` |
| `!=` | Not equal | `district != "Kathmandu"` |
| `>` / `<` | Greater / less | `age > 40` |
| `>=` / `<=` | Equal or more / less | `systolic >= 140` |
| `&` | AND — both true | `age > 40 & gender == "F"` |
| `\|` | OR — at least one true | `age < 25 \| age > 60` |
| `%in%` | Matches value in a list | `district %in% c("Lalitpur","Bhaktapur")` |

------------------------------------------------------------------------

### `mutate()` — Add New Columns

```{r}
# Add a single column with if_else()
patients %>%
  mutate(
    bp_status = if_else(systolic >= 140, "High", "Normal")
  )
```

```{r}
# Add multiple columns at once
patients %>%
  mutate(
    bp_status      = if_else(systolic >= 140, "High", "Normal"),
    glucose_status = if_else(glucose > 126, "Diabetic", "Normal"),
    senior         = if_else(age >= 60, TRUE, FALSE)
  )
```

------------------------------------------------------------------------

### `arrange()` — Sort Rows

```{r}
# Sort by glucose — highest first
patients %>% arrange(desc(glucose))
```

```{r}
# Sort by district then age within each district
patients %>% arrange(district, age)
```

------------------------------------------------------------------------

### 🔄 Your Turn 02 — select, filter, mutate, arrange *(5 minutes)*

> **Instructions:** Fill in column names and values. Column names do NOT need quotes. Text values like `"F"` or `"Kathmandu"` DO need quotes.

```{r}
# 1. Show name, age, and systolic for patients with systolic >= 130
#    Hint: select(name, age, ___) then filter(systolic >= 130)
patients %>%
  select( ,  , ) %>%
  filter( >= 130)
```

```{r}
# 2. Add column "risk_flag":
#    "At Risk" if systolic >= 140 OR (|) glucose > 126
#    "Normal" otherwise
patients %>%
  mutate(risk_flag = if_else(
     >= 140 |  > 126, "At Risk", "Normal"))
```

```{r}
# 3. Sort by glucose descending, show only the top 3 rows
#    Hint: arrange(desc(___)) then head(3)
patients %>%
  arrange(desc( )) %>%
  head( )
```

```{r}
# 4. ★ Female patients aged > 30, sorted by glucose (highest first)
#    Show only: name, age, glucose, district
patients %>%
  filter(gender ==  &  > 30) %>%
  arrange(desc( )) %>%
  select( ,  ,  , )
```

------------------------------------------------------------------------

### `group_by()` + `summarise()` — Calculate by Group

```{r}
# Average glucose — overall (one number)
mean(patients$glucose)
```

```{r}
# Average glucose — separately for each gender
patients %>%
  group_by(gender) %>%
  summarise(avg_glucose = mean(glucose))
```

```{r}
# Multiple statistics grouped by district
patients %>%
  group_by(district) %>%
  summarise(
    n_patients   = n(),
    avg_age      = round(mean(age), 1),
    avg_systolic = round(mean(systolic), 1),
    avg_glucose  = round(mean(glucose), 1),
    max_glucose  = max(glucose),
    pct_high_bp  = round(mean(systolic >= 140) * 100, 1)
  )
```

------------------------------------------------------------------------

### 🔄 Your Turn 03 — group_by + summarise *(5 minutes)*

> **Instructions:** Fill in the column name inside `group_by()` and inside each summary function. Use `n()` (no argument) to count rows, `mean(column)` for averages.

```{r}
# 1. Count how many patients are in each district
patients %>%
  group_by(___) %>%
  summarise(count = n())
```

```{r}
# 2. Average systolic and glucose by gender (round both to 1 decimal)
#    Hint: group_by(gender) → summarise(avg_systolic = round(mean(systolic), 1), ...)
patients %>%
  group_by(___) %>%
  summarise(
    avg_systolic = round(mean(___), 1),
    avg_glucose  = round(mean(___), 1)
  )
```

```{r}
# 3. ★ For each district: count patients, average glucose,
#    AND percentage with glucose > 126 (diabetic range)
#    Hint: mean(glucose > 126) gives proportion; multiply by 100 for %
patients %>%
  group_by(___) %>%
  summarise(
    n            = n(),
    avg_glucose  = mean(___),
    pct_diabetic = round(mean(___ > 126) * 100, 1)
  ) %>%
  arrange(desc(pct_diabetic))
```

------------------------------------------------------------------------

## PART 4 — Helper Functions

------------------------------------------------------------------------

```{r}
# distinct() — find unique values
patients %>% distinct(district)
patients %>% distinct(gender, district)
```

```{r}
# count() — quick frequency table (shortcut for group_by + n())
patients %>% count(district)
patients %>% count(district, sort = TRUE)   # sorted by frequency
```

```{r}
# slice() — pick rows by position or rank
patients %>% slice(1:3)                     # first 3 rows
patients %>% slice_max(glucose, n = 3)      # 3 highest glucose patients
patients %>% slice_min(systolic, n = 2)     # 2 lowest BP patients
patients %>% slice_sample(n = 4)            # 4 random rows
```

```{r}
# n_distinct() — count unique values
patients %>% summarise(n_districts = n_distinct(district))
patients %>%
  group_by(gender) %>%
  summarise(n_districts = n_distinct(district))
```

------------------------------------------------------------------------

### 🔄 Your Turn 04 — Helper Functions *(4 minutes)*

> **Instructions:** Fill in the column name or number in each blank.

```{r}
# 1. How many unique districts are in the dataset?
patients %>% distinct(___)
```

```{r}
# 2. Frequency table of patients by district, sorted by count (sort = TRUE)
patients %>% count(___, sort = ___)
```

```{r}
# 3. Show the 2 patients with the HIGHEST glucose
patients %>% slice_max(___, n = ___)
```

```{r}
# 4. Show the 3 patients with the LOWEST systolic BP
patients %>% slice_min(___, n = ___)
```

```{r}
# 5. ★ How many unique genders are in each district?
patients %>%
  group_by(___) %>%
  summarise(n_genders = n_distinct(___))
```

------------------------------------------------------------------------

## PART 5 — Working with Text: `stringr`

------------------------------------------------------------------------

```{r}
# str_detect — does the text contain a pattern? Returns TRUE/FALSE
patients %>% filter(str_detect(district, "pur"))   # Lalitpur, Bhaktapur
```

```{r}
# str_sub — extract characters by position
str_sub("Kathmandu", 1, 5)    # → "Kathm"
```

```{r}
# str_replace / str_remove
str_replace("High BP", "High", "Elevated")   # → "Elevated BP"
str_remove("P01_visit1", "_visit1")          # → "P01"
```

```{r}
# str_to_upper / str_to_lower / str_to_title
str_to_upper("anita")            # → "ANITA"
str_to_title("anita sharma")     # → "Anita Sharma"
```

```{r}
# str_trim — remove extra whitespace (common data entry problem)
str_trim("  Kathmandu  ")        # → "Kathmandu"
```

```{r}
# In mutate — create a district code from first 3 letters
patients %>%
  mutate(district_code = str_sub(district, 1, 3) %>% str_to_upper()) %>%
  select(name, district, district_code)
```

------------------------------------------------------------------------

### 🔄 Your Turn 05 — stringr *(4 minutes)*

> **Instructions:** Fill in the blanks. For `str_to_upper()`, put the column name inside. For `str_replace()`, the format is `str_replace(column, "old_text", "new_text")`.

```{r}

# Load clinical notes dataset
notes <- read_csv("data/notes.csv", show_col_types = FALSE)
notes
```

```{r}

# 1. Filter notes that mention "HIGH" (case sensitive)
notes %>%
  filter(str_detect(note, "HIGH"))
```

```{r}

# 2. Convert all notes to UPPERCASE for consistency
#    Hint: str_to_upper(COLUMN_NAME)
notes %>%
  mutate(note_clean = str_to_upper(___))
```

```{r}

# 3. Replace "HIGH" with "ELEVATED" in all notes
#    Hint: str_replace(note, "HIGH", "ELEVATED")
notes %>%
  mutate(note = str_replace(note, "HIGH", "___"))
```

```{r}

# 4. ★ Extract the systolic number from each note
#    Hint: str_extract(note, "[0-9]+") — matches the FIRST number found
notes %>%
  mutate(systolic_text = str_extract(note, "___"))
```

------------------------------------------------------------------------

## PART 6 — Conditional Labeling

------------------------------------------------------------------------

```{r}

# if_else() — exactly two categories
patients %>%
  mutate(bp_flag = if_else(systolic >= 140, "High", "Normal"))
```

```{r}
# case_when() — three or more categories
# The ~ (tilde) means "then"
patients %>%
  mutate(
    bp_category = case_when(
      systolic < 120                   ~ "Normal",
      systolic >= 120 & systolic < 130 ~ "Elevated",
      systolic >= 130 & systolic < 140 ~ "High Stage 1",
      systolic >= 140                  ~ "High Stage 2"
    )
  )
```

> 💡 `case_when()` checks conditions from top to bottom. First match wins. `TRUE ~` is the catch-all.

------------------------------------------------------------------------

### 🔄 Your Turn 06 — case_when() *(5 minutes)*

> **Instructions:** Fill in the threshold numbers. Use the clinical reference ranges from the top of this document.

```{r}
# 1. Add "glucose_category" with 3 levels
#    Normal: below 100 · Pre-diabetic: 100 to 125 · Diabetic: 126 and above
patients %>%
  mutate(
    glucose_category = case_when(
      glucose < ___                      ~ "Normal",
      glucose >= ___ & glucose <= ___    ~ "Pre-diabetic",
      glucose >= ___                     ~ "Diabetic"
    )
  )
```

```{r}
# 2. Add "age_group" with 3 levels
#    Young Adult: below 35 · Middle-aged: 35 to 55 · Senior: above 55
patients %>%
  mutate(age_group = case_when(
    age < ___    ~ "Young Adult",
    age <= ___   ~ "Middle-aged",
    TRUE         ~ "___"           # TRUE = catch-all for anything that didn't match
  ))
```

```{r}
# 3. ★ Add "combined_risk" combining BOTH BP and glucose
#    Dual Risk: systolic >= 140 AND glucose > 126
#    High BP Only: systolic >= 140 only
#    High Glucose Only: glucose > 126 only
#    Normal: neither
#    Then count how many patients fall into each group
patients %>%
  mutate(
    combined_risk = case_when(
      systolic >= ___ & glucose > ___ ~ "Dual Risk",
      systolic >= ___                 ~ "High BP Only",
      glucose > ___                   ~ "High Glucose Only",
      TRUE                            ~ "Normal"
    )
  ) %>%
  count(___)
```

------------------------------------------------------------------------

## PART 7 — Joining Datasets

------------------------------------------------------------------------

> In real health work, data lives in separate tables connected by a shared column (like patient ID).

```{r}
patients_basic <- read_csv("data/patients_basic.csv", show_col_types = FALSE)
patients_basic
```

```{r}
lab_results <- read_csv("data/lab_results.csv", show_col_types = FALSE)
lab_results
```

```{r}
# left_join — ALL patients, add results where available (missing = NA)
# Most common join in health data
left_join(patients_basic, lab_results, by = "patient_id")
```

```{r}
# inner_join — ONLY patients who have both a record AND results
inner_join(patients_basic, lab_results, by = "patient_id")
```

```{r}
# anti_join — patients with NO results yet
anti_join(patients_basic, lab_results, by = "patient_id")
```

```         
patients_basic: P01  P02  P03  P04  P05
lab_results:    P01  P02       P04       P06

left_join  → P01  P02  P03  P04  P05     (all from left, NA where missing)
inner_join → P01  P02       P04           (matched only)
anti_join  →           P03       P05     (no match in right)
```

------------------------------------------------------------------------

### 🔄 Your Turn 07 — Joins *(5 minutes)*

> **Instructions:** Fill in the join function name or the column reference in each blank.

```{r}
# 1. Keep ALL patients, attach results where available
#    Hint: left_join keeps all rows from the LEFT (first) table
___(patients_basic, lab_results, by = "patient_id")
```

```{r}
# 2. Keep ONLY patients who have both a record AND lab results
#    Hint: inner_join — matched rows only
___(patients_basic, lab_results, by = "patient_id")
```

```{r}
# 3. Which patients have NOT been tested yet?
#    Hint: anti_join — left rows with NO match in right
___(patients_basic, lab_results, by = "patient_id")
```

```{r}
# 4. ★ Left join, then filter to show only patients where cholesterol is NA
#    Hint: filter(is.na(COLUMN_NAME))
left_join(patients_basic, lab_results, by = "patient_id") %>%
  filter(is.na(___))
```

```{r}
# 5. ★★ Full join, add a column "tested" = TRUE if hemoglobin is NOT missing
#    Hint: !is.na(hemoglobin) — the ! means "NOT"
full_join(patients_basic, lab_results, by = "patient_id") %>%
  mutate(tested = !is.na(___))
```

------------------------------------------------------------------------

## PART 8 — Tidy Data with `tidyr`

------------------------------------------------------------------------

**Three rules of tidy data:**

1.  Each **variable** has its own column
2.  Each **observation** has its own row
3.  Each **cell** holds one value

```{r}
# Load blood pressure data — currently in WIDE format
bp_wide <- read_csv("data/bp_wide.csv", show_col_types = FALSE)
bp_wide    # one column per visit
```

```{r}
# Convert to LONG (tidy) format using pivot_longer()
bp_long <- bp_wide %>%
  pivot_longer(
    cols      = starts_with("visit"),   # which columns to collapse
    names_to  = "visit",                # new column: old column names go here
    values_to = "systolic"              # new column: the values go here
  )

bp_long    # now one row per patient per visit
```

```{r}
# Convert BACK to wide format using pivot_wider()
bp_long %>%
  pivot_wider(
    names_from  = "visit",     # column whose values become column names
    values_from = "systolic"   # column whose values fill the cells
  )
```

------------------------------------------------------------------------

### 🔄 Your Turn 08 — Pivoting *(6 minutes)*

> **Instructions:** For `pivot_longer()`, fill in: `cols` (which columns to collapse), `names_to` (name for the new column of old names), `values_to` (name for the new column of values). Use quotes around text strings in `names_to` and `values_to`.

```{r}
# Load glucose monitoring data — wide format (one column per month)
glucose_monitoring <- read_csv("data/glucose_monitoring.csv", show_col_types = FALSE)
glucose_monitoring
```

```{r}
# 1. Convert to LONG format
#    We want 3 columns: patient | month | glucose
#    Hint: cols = starts_with("month"), names_to = "month", values_to = "glucose"
glucose_long <- glucose_monitoring %>%
  pivot_longer(
    cols      = starts_with("___"),   # collapse all month_* columns
    names_to  = "___",                # new column: will hold "month_1", "month_2"...
    values_to = "___"                 # new column: will hold the glucose values
  )

glucose_long   # check: should have 16 rows (4 patients × 4 months)
```

```{r}
# 2. Average glucose per patient (across all 4 months)
#    Hint: group_by(patient) then summarise(avg_glucose = mean(glucose))
glucose_long %>%
  group_by(___) %>%
  summarise(avg_glucose = mean(___))
```

```{r}
# 3. ★ Which patient improved the MOST from month_1 to month_4?
#    Hint: filter month_1 and month_4, pivot_wider, mutate(improvement = month_1 - month_4)
glucose_long %>%
  filter(month %in% c("___", "___")) %>%
  pivot_wider(names_from = "___", values_from = "___") %>%
  mutate(improvement = month_1 - month_4) %>%
  arrange(desc(improvement))
```

------------------------------------------------------------------------

## PART 9 — End-to-End Pipeline

------------------------------------------------------------------------

### 🔄 Your Turn 09 — Full Pipeline *(10 minutes)*

> **Context:** A district health office needs to compare disease burden using **incidence rate per 100,000 people** (cases ÷ population × 100,000), which adjusts for different population sizes.
>
> **Instructions:** Build the pipeline one step at a time. Run each step, check the output, then add the next pipe.

```{r}
disease_data <- read_csv("data/disease_data.csv", show_col_types = FALSE)
disease_data
```

```{r}
# Step 1: Filter — keep only year 2022 data
#    Hint: filter(year == 2022)
disease_data %>%
  filter(year == ___)
```

```{r}
# Step 2: Add to Step 1 — remove the "disease" column
#    Hint: select(-disease)
disease_data %>%
  filter(year == 2022) %>%
  select(-___)
```

```{r}
# Step 3: Add to Step 2 — calculate rate per 100,000 people
#    Formula: (cases / population) * 100000, rounded to 2 decimal places
disease_data %>%
  filter(year == 2022) %>%
  select(-disease) %>%
  mutate(rate_per_100k = round((___ / ___) * 100000, 2))
```

```{r}
# Step 4: Add to Step 3 — what is the AVERAGE rate across all 2022 districts?
#    Hint: summarise(avg_rate = mean(rate_per_100k))
disease_data %>%
  filter(year == 2022) %>%
  select(-disease) %>%
  mutate(rate_per_100k = round((cases / population) * 100000, 2)) %>%
  summarise(avg_rate = mean(___))
```

```{r}
# Step 5: ★ Which district has the HIGHEST rate in 2022?
#    Hint: arrange(desc(rate_per_100k)) after the mutate step
disease_data %>%
  filter(year == 2022) %>%
  mutate(rate_per_100k = round((cases / population) * 100000, 2)) %>%
  arrange(desc(___))
```

------------------------------------------------------------------------

## 🧬 Case Study — Introduction to Genomics Data

------------------------------------------------------------------------

> *"Everything you practiced today with patient data applies directly to genomics data. The functions are the same — only the biology changes."*

------------------------------------------------------------------------

### Key Terms

| Term | What It Means | Health Analogy |
|------------------------|------------------------|------------------------|
| **Gene** | Segment of DNA producing a protein | A specific blood test |
| **Expression** | How actively a gene is used | The test result value |
| **RNA-seq** | Technique counting all active genes in a sample | Running all blood tests at once |
| **Count / CPM** | Normalized times a gene was detected | The measurement reading |
| **Sample** | One biological specimen | One patient visit |
| **Condition** | Experimental group | Control vs Treatment |
| **log2(Count + 1)** | Log-scale transformation | Compresses huge value ranges |
| **Fold change** | Treated ÷ Control | "2× higher" or "4× lower" |

> **Why `log2(Count + 1)`?** Gene counts range from 0 to 100,000+. Log2 makes all genes visible in one plot. A difference of 1 on log2 = 2× more expression. The `+1` prevents `log2(0) = -Inf`.

------------------------------------------------------------------------

### The Count Matrix Structure

```{r}
rnaseq_wide <- read_csv("data/rnaseq_counts_raw.csv", show_col_types = FALSE)
rnaseq_wide    # wide format: one column per sample
```

```{r}
# Reshape to long format — same pivot_longer() you just practiced!
rnaseq_long <- rnaseq_wide %>%
  pivot_longer(
    cols      = starts_with("Control") | starts_with("Treated"),
    names_to  = "sample",
    values_to = "count"
  )

rnaseq_long
```

```{r}
# Load sample metadata and join — same left_join() you just practiced!
sample_meta <- read_csv("data/sample_meta.csv", show_col_types = FALSE)

rnaseq_full <- left_join(rnaseq_long, sample_meta, by = "sample")
rnaseq_full
```

------------------------------------------------------------------------

### 🔄 Case Study Exercise — RNA-seq Manipulation *(10 minutes)*

> **Context:** RNA-seq experiment comparing **Control** vs **Drug-treated** cells. Apply Day 3 skills to real genomics data.
>
> **Instructions:** Fill in the blanks. Steps 1–4 use functions you already know from today.

```{r}
counts_raw  <- read_csv("data/rnaseq_counts_raw.csv", show_col_types = FALSE)
sample_meta <- read_csv("data/sample_meta.csv", show_col_types = FALSE)

counts_raw
sample_meta
```

```{r}
# Step 1: Reshape to long format
#    We want 3 columns: gene_name | sample | cpm
#    Hint: cols = starts_with("Control") | starts_with("Treated")
counts_long <- counts_raw %>%
  pivot_longer(
    cols      = starts_with("Control") | starts_with("___"),
    names_to  = "___",
    values_to = "___"
  )

counts_long
```

```{r}
# Step 2: Join with sample metadata
#    Hint: left_join(counts_long, sample_meta, by = "sample")
counts_full <- left_join(___, ___, by = "___")
counts_full
```

```{r}
# Step 3: Average CPM per gene per condition
#    Hint: group_by(gene_name, condition) then summarise(avg_cpm = round(mean(cpm), 1))
avg_expression <- counts_full %>%
  group_by(___, ___) %>%
  summarise(avg_cpm = round(mean(___), 1), .groups = "drop")

avg_expression
```

```{r}
# Step 4: Which 3 genes are most highly expressed in each condition?
#    Hint: group_by(condition) then slice_max(avg_cpm, n = 3)
avg_expression %>%
  group_by(___) %>%
  slice_max(___, n = 3)
```

```{r}
# Step 5: ★ Calculate fold change (Treated ÷ Control) and log2 fold change
#    Hint: pivot_wider(names_from = condition, values_from = avg_cpm)
#    Then: mutate(fold_change = Treated / Control)
#          mutate(log2_fc = log2(Treated / Control))
#          mutate(direction = case_when(log2_fc > 1 ~ "Up-regulated",
#                                       log2_fc < -1 ~ "Down-regulated",
#                                       TRUE ~ "No change"))
avg_expression %>%
  pivot_wider(names_from = ___, values_from = ___) %>%
  mutate(
    fold_change = round(Treated / ___, 2),
    log2_fc     = round(log2(Treated / ___), 2),
    direction   = case_when(
      log2_fc > ___  ~ "Up-regulated",
      log2_fc < ___  ~ "Down-regulated",
      TRUE           ~ "No change"
    )
  ) %>%
  arrange(desc(abs(log2_fc)))
```

> 🔍 Which genes are up-regulated by the drug? VEGF and HIF1A going up together is a known hypoxia/tumor pattern in biology.

------------------------------------------------------------------------

## DAY 3 RECAP

| Skill          | Key Functions                                              |
|------------------------------------|------------------------------------|
| Load data      | `read_csv()`, `write_csv()`, `read_excel()`                |
| Missing values | `is.na()`, `drop_na()`, `replace_na()`, `na.rm = TRUE`     |
| Pipe           | `%>%`                                                      |
| Choose columns | `select()`, `starts_with()`, `ends_with()`                 |
| Choose rows    | `filter()` + logical operators                             |
| Add columns    | `mutate()`, `if_else()`, `case_when()`                     |
| Sort           | `arrange()`, `arrange(desc(...))`                          |
| Helpers        | `count()`, `distinct()`, `slice_max()`, `n_distinct()`     |
| Text           | `str_detect()`, `str_replace()`, `str_sub()`, `str_trim()` |
| Summarize      | `group_by()` + `summarise()`                               |
| Joins          | `left_join()`, `inner_join()`, `anti_join()`               |
| Reshape        | `pivot_longer()`, `pivot_wider()`                          |

------------------------------------------------------------------------

## 📎 Quick Reference

```{r}
#| eval: false

# ── PIPE ──────────────────────────────
df %>% step1() %>% step2() %>% step3()

# ── dplyr ─────────────────────────────
select(col1, col2)          # keep columns
select(-col3)               # remove column
filter(age > 40)
filter(district %in% c("A","B"))
mutate(new_col = old_col * 2)
arrange(desc(glucose))
group_by(district) %>%
  summarise(n = n(), avg = mean(value))
if_else(condition, "Yes", "No")
case_when(x < 100 ~ "Normal",
          x <= 125 ~ "Pre-diabetic",
          TRUE ~ "Diabetic")
count(district, sort = TRUE)
distinct(district)
slice_max(glucose, n = 3)

# ── MISSING VALUES ────────────────────
colSums(is.na(df))
mean(df$col, na.rm = TRUE)
df %>% drop_na()
df %>% drop_na(column)
df %>% mutate(col = replace_na(col, 0))

# ── JOINS ─────────────────────────────
left_join(df1, df2, by = "id")
inner_join(df1, df2, by = "id")
anti_join(df1, df2, by = "id")

# ── RESHAPE ───────────────────────────
pivot_longer(cols = starts_with("visit"),
             names_to = "visit",
             values_to = "value")
pivot_wider(names_from = "visit",
            values_from = "value")
```
